---
title: "Tidyverse"
output: learnr::tutorial
runtime: shiny_prerendered
---

<!-- EF: copied Jessica Edward's Tutorial Template we discussed in meeting. Thank you Jessica! --> 

```{r setup, include=FALSE}
library(tidyverse)
library(learnr)
knitr::opts_chunk$set(echo = FALSE)

# I've commented this out so that it doesn't appear in the bok repo

# options(tutorial.storage = filesystem_storage(dir = "./storage", compress = TRUE))

library(gapminder)
gapminder_unfiltered


```

### Question 1

Load the gapminder library: 

```{r exercise-1, exercise=TRUE}


```

```{r exercise-1-solution}
library(gapminder)

```

### Question 2
`glimpse()` the `gapminder_unfiltered` dataset:
```{r exercise-2, exercise=TRUE}

```

```{r exercise-2-solution}
glimpse(gapminder_unfiltered)
```

### Question 3
Take a `summary()` of the `gapminder_unfiltered` dataset:
```{r exercise-3, exercise=TRUE}

```

```{r exercise-3-solution}
summary(gapminder_unfiltered)
```

### Question 4
`View()` the `gapminder_unfiltered` dataset:
```{r exercise-4, exercise=TRUE}

```

```{r exercise-4-solution}
View(gapminder_unfiltered)
```

```{r quiz-4}
quiz(
  question("Which of these is a column within the dataframe",
    answer("age"),
    answer("gdp"),
    answer("lifeExp", correct = TRUE),
    answer("month")
  )
)
```

### Question 5
Use the extraction operator `$` to extract the `country` column in the `gapminder_unfiltered` dataset:
```{r exercise-5, exercise = TRUE}

```

```{r exercise-5-solution}
gapminder_unfiltered$country
```

```{r quiz-5}
quiz(
  question("How many levels are there?",
    answer("202"),
    answer("43"),
    answer("189"),
    answer("187", correct = TRUE)
  )
)
```

### Question 6
Do these same as in Question 5 but with the `pull()` function. 
```{r exercise-6, exercise=TRUE}

```

```{r exercise-6-solution}
pull(.data = gapminder_unfiltered, var = country)
```

### Question 7
`slice()` the first 5 lines of the `gapminder_unfiltered` dataset
```{r exercise-7, exercise=TRUE}

```

```{r exercise-7-solution}
slice(.data = gapminder_unfiltered, 1:5)
```

### Question 8
Use the help page to find out the arguments for `dplyr::filter`
```{r exercise-8, exercise=TRUE}

```

```{r exercise-8-solution}
help(topic = filter, package = dplyr)
# Run the following `?dplyr::filter` in the console 

```

### Question 9
Use `filter()` to pick out all of the entries (rows) in the dataset where the country is Finland in the `gapminder_unfiltered` dataset. Do not use the pipe `%>%` operator.
```{r exercise-9, exercise = TRUE}

```

```{r exercise-9-solution}
filter(.data = gapminder_unfiltered, country == "Finland")
```

### Question 10
Do the same as in question 9 but use the pipe `%>%` operator.
```{r exercise-10, exercise = TRUE}

```

```{r, exercise-10-solution}
gapminder_unfiltered %>%
  filter(country == "Finland")
```

### Question 11
Create a variable `gap_filter` that filters observations for years in the `year_list` from the `gapminder_unfiltered` dataset. Hint: use the `%in%` operator. 

```{r exercise-11, exercise=TRUE}
year_list <- c(1952, 1957, 1962, 1967, 1972, 1977, 1982, 1987, 1992, 1997, 2002, 2007)


```

```{r exercise-11-solution}
gap_filter <- gapminder_unfiltered %>%
  filter(year %in% year_list)
```

### Question 12 
Use the `mutate()` function to create a variable `gdp` that is the product of `pop` and `gdpPercap` from the `gap_filter` dataframe created in the previous question. Don't forget to reassign this mutated dataframe as `gap_filter`.
```{r exercise-12, exercise = TRUE}

```

```{r exercise-12-solution}
gap_filter <- gap_filter %>%
  mutate(gdp = pop * gdpPercap)
```

### Question 13
Use the `filter()` function to pick out the observation from Thailand in 1997. Name this observation `thailand_97`
```{r exercise-13, exercise = TRUE}

```

```{r exercise-13-solution}
thailand_97 <- gap_filter %>%
  filter(country == "Thailand",
         year == 1997)
```

### Exercise 14
`pull()` the `gdp` variable from the `thailand_97` variable.
```{r exercise-14, exercise=TRUE}

```

```{r exercise-14-solution}
pull(.data = thailand_97, var = gdp)
```

### Question 15
Create a variable `top_gdp` and `arrange()` the `gap_filter` dataframe from the highest to the lowest gdp. Note, disregard the `thailand_97` variable from the previous 2 questions.
```{r exercise-15, exercise=TRUE}

```

```{r exercise-15-solution}
top_gdp <- gap_filter %>%
  arrange(desc(gdp))
```

### Question 16
`slice()` the top 5 rows from the `top_gdp` dataframe.
```{r exercise-16, exercise=TRUE}

```

```{r exercise-16-solution}
slice(.data = top_gdp, 1:5)
```

### Question 17
Use the `top_n()` function to select the top 5 observations corresponding to the highest `gdp` in the `gap_filter` dataframe. How is the result different from the result in Question 16? Input `?top_n` in the console if you need a reminder about how to use `top_n()`.
```{r exercise-17, exercise = TRUE}

```

```{r exercise-17-solution}
top_n(x = gap_filter, n = 5, wt = gdp)
```

### Question 18
`select()` the variables `country`, `year`, `gdp`, and `gdpPercap` from the `gap_filter` dataframe. Name this new dataframe `gap_subset`.
```{r exercise-18, exercise=TRUE}

```

```{r exercise-18-solution}
gap_subset <- gap_filter %>%
  select(country, year, gdp, gdpPercap)
```

### Question 19
Take the `gap_filter` dataset and `group_by()` `year`. Rename this dataframe `gap_filter`.
```{r exercise-19, exercise=TRUE}

```

```{r exercise-19-solution}
gap_filter <- gap_filter %>%
  group_by(year)
```

### Exercise 20
`summarize()` the `gap_filter` dataframe and find the `mean()` `gdp` and the `mean()` `gdpPercap` for each year. Name these new variable `mean_gdp` and `mean_gdpPercap`, respectively.
```{r exercise-20, exercise=TRUE}

```

```{r exercise-20-solution}
gap_filter %>%
  summarize(mean_gdp = mean(gdp),
         mean_gdpPercap = mean(gdpPercap))
```

## Topic 2 - factors

### Exercise 21
Load the `fueleconomy` package with the `library()` function. If you have not downloaded the fueleconomy library, yet, run `install.packages("fueleconomy")` in your console.
```{r exercise-21, include = FALSE}

```

```{r exercise-21-solution}
library(fueleconomy)
```

### Question 22
Run the following: `help(package = "fueleconomy")` in your console.
```{r quiz-22}
quiz(
  question("Which of these dataframes is not in the fueleconomy package",
    answer("mileage", correct = TRUE),
    answer("common"),
    answer("vehicles")
  )
)
```

### Question 23
`glimpse()` the `vehicles` dataset:
```{r exercise-23, exercise=TRUE}

```

```{r exercise-23-solution}
glimpse(vehicles)
```

### Question 24
Take a `summary()` of the `vehicles` dataset:
```{r exercise-24, exercise=TRUE}

```

```{r exercise-24-solution}
summary(vehicles)
```

### Question 25
`View()` the `vehicles` dataset:
```{r exercise-25, exercise=TRUE}

```

```{r exercise-25-solution}
View(vehicles)
```

### Question 26
Make the `class` column in the `vehicles` dataframe a factor instead of a character `chr` variable. Do this by using the `mutate()` and `factor()` functions. Reassign this changed dataframe as `vehicles`. 
```{r exercise-26, exercise = TRUE}

```

```{r exercise-26-solution}
vehicles <- vehicles %>%
  mutate(class = factor(class))
```

### Question 27
Create a dataframe `vehicles_grouped` that uses the `group_by()` function for the `class` variable. 
```{r exercise-27, exercise = TRUE}

```

```{r exercise-27-solution}
vehicles_grouped <- vehicles %>%
  group_by(class)
```

### Question 28
Create a `mean_cty` variable using the `mutate()` and `mean()` functions for the `vehicles_grouped` dataframe. Reassign this mutated dataframe as vehicles grouped.
```{r exercise-28, exercise = TRUE}

```

```{r exercise-28-solution}
vehicles_grouped <- vehicles_grouped %>%
  summarize(mean_cty = mean(cty))
```

### Question 29
Create a ggplot with the independent variable as `class`, the dependent variable as `mean_cty`, and the `geom_point()` function. Names this plot `vehicles_plot`.
```{r exercise-29, exercise = TRUE}

```

```{r exercise-29-solution}
vehicles_plot <- vehicles_grouped %>% 
  ggplot(aes(x = class, y = mean_cty)) +
  geom_point()
```

### Question 30
Flip the coordinates of the `vehicles_plot` graphic. Rename the flipped graphic `vehicles_plot`.
```{r exercise-30, exercise = TRUE}

```

```{r exercise-30-hint}
`Look at the coord_flip() function`
```

```{r exercise-30-solution}
vehicles_plot <- vehicles_plot +
  coord_flip()
```

### Question 31 - CHALLENGE
One of the key skills that this course aims to teach you is problem solving. As you notice, in the previous graphic, the levels of the `class` variable are not in a preferred order. The ability to search for and implement the solutions to such problems will be a tenet of you data analytics experience. To help you along, here's a [link](https://rstudio-pubs-static.s3.amazonaws.com/7433_4537ea5073dc4162950abb715f513469.html). Attempt to reorder `vehicles_grouped$class` by `vehicles_grouped$mean_cty`.

```{r exercise-31, exercise = TRUE}

```

```{r exercise-31-solution}
vehicles_grouped$class <- factor(vehicles_grouped$class, 
                                 levels = vehicles_grouped$class[order(vehicles_grouped$mean_cty)] )
```

### Question 32
Recreate the flipped ggplot with `class` as the independent variable and `mean_cty` as the dependent variable. Assign this plot `vehicles_plot_2`. 
```{r exercise-32, exercise = TRUE}

```

```{r exercise-32-solution}
vehicles_plot_2 <- vehicles_grouped %>%
  ggplot(aes(x = class, y = mean_cty)) +
  geom_point() +
  coord_flip()
```

### Question 33
Load the `ggthemes` package. Install it first if you haven't yet installed it.
```{r exercise-33, exercise = TRUE}

```

```{r exercise-33-solution}
library(ggthemes)
```

### Question 34
Apply the classic theme `theme_classic()` to `vehicles_plot_2`. Reassign the plot to `vehicles_plot_2`
```{r exercise-34, exercise = TRUE}

```

```{r exercise-34-solution}
vehicles_plot_2 <- vehicles_plot_2 +
  theme_classic()
```

### Question 35
Add a title and axis titles to `vehicles_plot_2` and reassign the graphic to `vehicles_plot_2`.
```{r exercise-35, exercise = TRUE}

```

```{r exercise-35-solution}
vehicles_plot_2 <- vehicles_plot_2 +
  labs(title = "Mean Mileage per Gallon (in the City) According to Vehicle Class",
       x = "Mean MPG",
       y = "Vehicle Class")
```

### Question 36
Load the `nycflights13` library. Don't forget to install the package if you havent already
```{r exercise-36, exercise = TRUE}

```

```{r exercise-36-solution}
library(nycflights13)
```

### Question 37
Run `help(package = nycflights13)` in your console
```{r quiz-37}
quiz(
  question("Which of these is not a dataset within the package",
    answer("flights"),
    answer("engines", correct = TRUE),
    answer("weather"),
    answer("planes")
  )
)
```

### Question 38
`glimpse()` the `flights` dataset:
```{r exercise-38, exercise=TRUE}

```

```{r exercise-38-solution}
glimpse(flights)
```

### Question 39
Take a `summary()` of the `flights` dataset:
```{r exercise-39, exercise=TRUE}

```

```{r exercise-39-solution}
summary(flights)
```

### Question 40
`View()` the `flights` dataset:
```{r exercise-40, exercise=TRUE}

```

```{r exercise-40-solution}
View(flights)
```

### Question 41
`glimpse()` the `planes` dataset:
```{r exercise-41, exercise=TRUE}

```

```{r exercise-41-solution}
glimpse(planes)
```

### Question 42
Take a `summary()` of the `planes` dataset:
```{r exercise-42, exercise=TRUE}

```

```{r exercise-42-solution}
summary(planes)
```

### Question 43
`View()` the `planes` dataset:
```{r exercise-43, exercise=TRUE}

```

```{r exercise-43-solution}
View(planes)
```

### Question 44
Take a subset of the `flights` dataset using the `select()` function. Pick out the variables `year`, `month`, `day`, `carrier`, `flight`, `tailnum`, `origin`, `dest`, `air_time` and `distance`. Name this new dataframe `flights_subset`.
```{r exercise-44, exercise = TRUE}

```

```{r exercise-44-solution}
flights_subset <- select(.data = flights, 
                         year, 
                         month, 
                         day, 
                         carrier, 
                         flight, 
                         tailnum, 
                         origin, 
                         dest, 
                         air_time, 
                         distance)
```

### Question 45
`left_join()` the `planes` dataframe and the `flights` dataframe. It will be up to you to decide what variable by which to join the two tibbles. Name the new tibble `planes_joined`
```{r exercise-45, exercise = TRUE}

```

```{r exercise-45-solution}
planes_joined <- left_join(x = planes, y = flights_subset, by = "tailnum")
```

### Question 46
`View()` the `planes_joined` dataset.
```{r exercise-46, exercise = TRUE}

```

```{r exercise-46-solution}
View(planes_joined)
```

### Question 47
`rename()` the `year.x` variable as `year_plane` and the `year.y` variable as `year_flight` in the `planes_joined` dataframe. Reassign this dataframe with the new variable names `planes_joined`.
```{r exercise-47, exercise = TRUE}

```

```{r exercise-47-solution}
planes_joined <- planes_joined %>%
  rename(year_plane = "year.x", year_flight = "year.y")
```

### Question 48
Use the `factor()` function to make the `model` variable in the `planes_joined` dataset a factor and not a character column.
```{r question-48, exercise = TRUE}

```

```{r question-48-solution}
planes_joined <- planes_joined %>%
  mutate(model = factor(model))
```

### Question 49
Run the `levels()` function on the `planes_joined$model` variable. 
```{r question-49, exercise = TRUE}

```

```{r question-49-solution}
levels(planes_joined$model)
```

```{r quiz-49}
quiz(
  question("How many levels are there?",
    answer("126"),
    answer("127", correct = TRUE),
    answer("80"),
    answer("143")
  )
)
```

### Question 50

```{r question-50, exercise = TRUE}

```

```{r question-50-solution}
planes_joined <- planes_joined %>%
  mutate(model = model %>% fct_infreq())
```

### Question 51
Use `group_by()` to group the `planes_joined` dataframe by model. Reassign this new dataframe as `planes_joined`.
```{r question-51, exercise = TRUE}

```

```{r question-51-solution}
planes_joined <- planes_joined %>%
  group_by(model)
```
### Question 52
Use the `count()` function to count the number of flights for each model of airplane. Reassign this new dataframe as `planes_joined` 
```{r question-52, exercise = TRUE}

```

```{r question-52-solution}
planes_joined <- planes_joined %>%
  count()
```

### Question 53
`slice()` the first 10 observation of the `planes_joined` tibble. Note, you will need to pass `planes_joined` as a `data.frame()` within the slice. Name the result `planes_10`
```{r question-53, exercise = TRUE}

```

```{r question-53-solution}
planes_10 <- slice(data.frame(planes_joined), 1:10)
```

### Question 54
Create a `ggplot` with the `planes_10` data. Let `model` be the independent variable, and let `n` be the dependent variable. Use the `geom_col()` function and assign this plot as `top_models`.
```{r question-54, exercise = TRUE}

```

```{r question-54-solution}
top_models <- ggplot(data = planes_10, aes(x = model, y = n)) +
  geom_col()
```

### Question 55 
Flip the coordinates of the plot. Use the `ggthemes` `theme_classic()` and give the plot a title and axis titles.
```{r question-55, exercise = TRUE}

```

```{r question-55-solution}
top_models + 
  coord_flip() + 
  theme_classic() + 
  labs(title = "Top Ten Plane Models by Number of Flights",
       x = "Model",
       y = "Number of Flights")
```

### Question 56
Use the `read_csv()` to read the harvard-faculty-gender.csv file. Assign this file to `gender_data`. You will need to set this URL https://raw.githubusercontent.com/davidkane9/bok/master/02-tidyverse/data/harvard-faculty-gender-final.csv as the file argument.
```{r question-56, exercise = TRUE}

```

```{r question-56-solution}
gender_data <- read_csv(file = "https://raw.githubusercontent.com/davidkane9/bok/master/02-tidyverse/data/harvard-faculty-gender-final.csv")
```

### Question 57
Run summary functions to familiarize yourself with the data. 
```{r question-57, exercise = TRUE}

```

```{r question-57-solution}
glimpse(gender_data)
summary(gender_data)
View(gender_data)
```

### Question 58
Repeat question 56, but this time, adjust your `argument` to skip the first row. In addition, `rename()` the `prof_of_practic_f` variable to `prof_of_practice_f`. Reassign this file to `gender_data`. 
```{r question-58, exercise = TRUE}

```

```{r question-58-solution}
gender_data <- read_csv(file = "https://raw.githubusercontent.com/davidkane9/bok/master/02-tidyverse/data/harvard-faculty-gender-final.csv", skip = 1) %>%
  rename(prof_of_practice_f = "prof_of_practic_f")
```

### Question 59
Repeat the summary functions Question 57
```{r question-59, exercise = TRUE}

```

```{r question-59-solution}
glimpse(gender_data)
summary(gender_data)
View(gender_data)
```

### Question 60
Run the `tail()` function on the `gender_data` dataset. 
```{r question-60, exercise = TRUE}

```

```{r question-60-solution}
tail(gender_data)
```

### Question 61 
`filter()` out the final row of the dataset. You will need to pick an appropriate identifier. 
```{r question-61, exercise = TRUE}

```

```{r question-61-solution}
gender_data <- gender_data %>%
  filter(concentration != "Total")

`Note you could also filter(!is.na(division)), but this could be problematic if there are multipe rows with NA values`
```

### Question 62
`mutate()` the `gender_data` data set to include a column with the total number of `full_profs` and another with the total number of `non_tenured_profs.`
```{r question-62, exercise = TRUE}

```

```{r question-62-solution}
gender_data <- gender_data %>%
  mutate(full_profs = full_profs_m + 
                      full_profs_f,
         non_tenured_profs = assoc_m + 
                             assoc_f + 
                             asst_m + 
                             asst_f + 
                             lec_precep_adj_m +
                             lec_precep_adj_f +
                             prof_of_practice_m +
                             prof_of_practice_f)
```

### Question 63
`arrange()` the `gender_data` dataframe by the number of male (full) professors. Assign this code chunk to `top_male()`. Then, call `top_male`. 
```{r question-63, exercise = TRUE}

```

```{r question-63-solution}
top_male <- gender_data %>%
  arrange(desc(full_profs_m))
top_male
```

### Question 64
Repeat the previous question with the number of female (full) professors. Assign this to `top_female`.
```{r question-64, exercise = TRUE}

```

```{r question-64-solution}
top_female <- gender_data %>%
  arrange(desc(full_profs_f))

top_female
```

### Question 65
`mutate()` a column that calculate the percent of female full professors (out of total full professors) and call it `perc_full_profs`. Do the same with female non-tenured professors called `perc_non_tenured`.
```{r question-65, exercise = TRUE}

```

```{r question-65-solution}
gender_data <- gender_data %>%
  mutate(perc_full_profs = full_profs_f / full_profs,
         perc_non_tenured = (assoc_f + asst_f + lec_precep_adj_f + prof_of_practice_f) / non_tenured_profs)
```

### Question 66
`arrange()` the `gender_data` tibble by the percentage of female (full) professors. Use `desc()`. `slice()` the top five, and `select()` the `concentration` and `perc_full_profs`. Name it `top_perc_female`.
```{r question-66, exercise = TRUE}

```

```{r question-66-solution}
top_perc_female <- gender_data %>%
  arrange(desc(perc_full_profs)) %>%
  slice(1:5) %>%
  select(concentration, perc_full_profs)
```

### Question 67
Repeat the previous question for the non-tenured professors. Assign the result as `top_perc_female_2`. 
```{r question-67, exercise = TRUE}

```

```{r question-67-solution}
top_perc_female_2 <- gender_data %>%
  arrange(desc(perc_non_tenured)) %>%
  slice(1:5) %>%
  select(concentration, perc_non_tenured)
```

### Question 68
`select()` a subset of the `gender_data` dataset. Pick out `concentration`, `perc_full_profs`, and `division`. Name this subset `division_data`.
```{r question-68, exercise = TRUE}

```

```{r question-68-solution}
division_data <- gender_data %>%
  select(concentration, perc_full_profs, division)
```

### Question 69
Use `group_by()` to group division data by `division`. Reassign this to `division_data`.
```{r question-69, exercise = TRUE}

```

```{r question-69-solution}
division_data <- division_data %>%
  group_by(division)
```

### Question 70
Use `summarize()` to take the `mean()` of `perc_full_profs` according to `division`. 
```{r question-70, exercise = TRUE}

```

```{r question-70-solution}
division_data %>%
  summarize(mean_f_per_division = mean(perc_full_profs))
```

### Question 71
`select()` a subset of the `gender_data`, `concentration`, `perc_full_profs`, `perc_non_tenured`, and `division`. Reassign this data table as `gender_data`.
```{r question-71, exercise = TRUE}

```

```{r question-71-solution}
gender_data <- gender_data %>%
  select(concentration, perc_full_profs, perc_non_tenured, division)
```

### Question 72
Use `pivot_longer()` on the `perc_full_profs` and `perc_non_tenured` variables. Set the new names to `prof_type` and the new values to `perc_profs_f`. Reassign this data table as `gender_data`.
```{r question-72, exercise = TRUE}

```

```{r question-72-solution}
gender_data <- gender_data %>%
  pivot_longer(cols = starts_with("perc"), 
               names_to = "prof_type",
               values_to = "perc_profs_f")
```

### Question 73
Create a ggplot named `perc_prof_data` that takes `concentration` as the independent variable, `perc_profs_f` as the dependent variable, and the fill as `prof_type`. Then, use `geom_col` and set the position to `dodge`.
```{r question-73, exercise = TRUE}

```

```{r question-73-solution}
perc_prof_data <- ggplot(gender_data, aes(x = concentration, y = perc_profs_f, fill = prof_type)) +
  geom_col(position = "dodge")
```

### Question 74
Use `theme_classic()` and change the titles (x-axis, y-axis, legend, and above the graph) and adjust the legend text. 
```{r question-74, exercise = TRUE}

```

```{r question-74-solution}
perc_prof_data <- perc_prof_data +
  labs(title = "Percentage of Female Professors by Concentration at Harvard",
       x = "Concentration",
       y = "Percentage of Female Professors",
       color = "Professor Type") + 
  scale_fill_discrete(labels = c("Tenured Professors", "Non-tenured Professors")) +
  theme_classic()
  
```

### Question 75
Turn the x-axis text -90 degrees. 
```{r question-75, exercise = TRUE}

```

```{r question-75-solution}
perc_prof_data + 
  theme(axis.text.x = element_text(angle = -90))
  
```


