---
title: "Type S and M errors"
output:
  xaringan::moon_reader:
    lib_dir: libs
    
    nature:
      ratio: 16:10
      
      # make it so that incremental slides are not numbered in slide numbers
      
      countIncrementalSlides: false
    
    # setting seal = false gets rid of the automatic title slide. Set class: title-slide at the top of first slide to create your own title slide below
    
    seal: false
      
# create a new slide with ---      
      
---

```{r, include=FALSE}
knitr::opts_chunk$set(fig.width = 6, message = F, warning = F, 
                      comment = "", cache = F)

## flipbook package
source(file = "https://raw.githubusercontent.com/EvaMaeRey/little_flipbooks_library/master/xaringan_reveal_parentheses_balanced.R")

library(tidyverse)

## slide presenter package
library(xaringan)

library(PPBDS.data)
library(gganimate)
```

```{r, include = F}
# Little bit complicated here, recreates sim_plot function from retrodesign but customized
## Set statistical significance requirement (alpha / p < 0.05)
alpha = 0.05

# 5000 sims, infinite degrees of freedom

df = Inf
n.sims = 5000

# True effect

A = 0.5

# 95% quantiles

z <- qt(1 - alpha/2, df)

# Calculate power 

p.hi <- 1 - pt(z - A, df)
p.lo <- pt(-z - A, df)
power <- p.hi + p.lo

## Calculate Type S error prob
typeS <- p.lo/power

## 5000 sims of random generation for t-distribution
estimate <- A + rt(n.sims, df)

## Sample 1000 to add in some funky randomness
estimate <- sample(estimate, 1000)

## Build dataframe that we'll plug into our plots
dat <- data.frame(y = estimate) %>%

## Change shape depending on type of error
  mutate(pch = case_when(y < -1.96 ~ 16,
                           y > 1.96 ~ 17,
                           TRUE ~ 15),

## Need an x for geom_point
         x = seq_along(y),
  
## Change color depending on whether error or stat insignificant
         color = case_when(abs(y) < 1.96 ~ "grey",
                           TRUE ~ "black"),
         )
```

```{r p1, echo = F}

## plot time!

## fig1, scatterplot of randomly generated estimates
plot <- ggplot(dat, aes(x = x, y = y, group = y, col = color)) +
  guides(colour = "none") + 
  xlab("n") +

## Custom legend
  scale_shape_manual(values = c(16, 17, 15), 
                     name = " ", 
                     labels = c("Statistically Insignificant", 
                                "Type S Errors",
                                "Type M Errors")) +
  guides(shape = guide_legend(override.aes = list(colour = c("grey", "black", "black")))) + 
  scale_color_manual(values = c("black", "grey")) + 
  geom_point(aes(pch = as.factor(pch)), size = 3) 

## fig2, add in true estimand

plot2 <- plot +
  geom_hline(yintercept = A, linetype = "dashed", size = 1)

# fig3, add in conf_int. The formula was "= s * z" when I first looked at this
# work. But "s" was not defined. So, I just deleted it. Maybe it should be 2 or
# 1.96?

plot3 <- plot2 + 
  geom_hline(yintercept = z, size = 1) + 
  geom_hline(yintercept = -z, size = 1) 

# fig4, zoom in on Type M

plot4 <- plot3 +
  ylim(1.94, 4)

# fig5, zoom in on Type S

plot5 <- plot3 +
  ylim(-2.5, -1.94)
```

<!-- align center -->

.center[

# What do Type M and S errors look like?

```{r, echo = F}

# animation of estimates being generated

plot3 +
  transition_reveal(x)
```

]

---
## Definitions

-Type M, or magnitude errors, occur when the test statistic drastically exaggerates the true effect. If a test has a true effect of 2, and our sample has a mean effect size of 20, we will have a Type M error.


-Type S, or sign errors, occur when the test statistic is in the opposite direction of the true effect. If a test has a true effect size of 2, and our sample has a mean effect size of -2, we will have a Type S error. 


---
<!-- create a left column -->


.pull-left[

##### To demonstrate potential Type S and Type M errors, let's simulate 5000 random draws from a population with a true mean of 0.5 and SD of 1.

]

--

<!-- create a right column -->

.pull-right[

```{r, echo = FALSE}
plot
```

] 

---

.pull-left[

##### A dashed line can represent this true mean.

]

.pull-right[
```{r, echo = FALSE}
plot
```

] 

---


.pull-left[

##### A dashed line can represent this true mean.
]

.pull-right[

```{r, echo = FALSE}
plot2
```

] 
---


.pull-left[

##### Assuming normality, any results within the standard confidence interval of -1.96 to 1.96 won't be statistically significant. 
]

.pull-right[

```{r, echo = FALSE}
plot2
```

]
---


.pull-left[
##### Assuming normality, any results within the standard confidence interval of -1.96 to 1.96 won't be statistically significant. 
]

.pull-right[
```{r, echo = FALSE}
plot3
```

] 
---

.pull-left[
##### This leaves two types of results that will be statistically significant, but erroneous. Results above the upper bound of our confidence interval will be Type M errors.
]

.pull-right[
```{r, echo = FALSE}
plot3
```

] 
---

.pull-left[
##### This leaves two types of results that will be statistically significant, but erroneous. Results above the upper bound of our confidence interval will be Type M errors.
]

.pull-right[
```{r, echo = FALSE}
plot4
```

] 

---

.pull-left[

##### And results below the lower bound of our confidence interval will be Type S errors.

]

.pull-right[

```{r, echo = FALSE}
plot4
```
]
---
.pull-left[
##### And results below the lower bound of our confidence interval will be Type S errors.
] 

.pull-right[

```{r, echo = FALSE}
plot5
```
]
---


