---
title: "Type 1 Errors?"
output:
  xaringan::moon_reader:
    lib_dir: libs
    
    nature:
      ratio: 16:10
      
      # make it so that incremental slides are not numbered in slide numbers
      
      countIncrementalSlides: false
    
    # setting seal = false gets rid of the automatic title slide. Set class: title-slide at the top of first slide to create your own title slide below
    
    seal: false
      
# create a new slide with ---      
      
---

```{r, include=FALSE}
knitr::opts_chunk$set(fig.width = 6, message = F, warning = F, 
                      comment = "", cache = F)


library(tidyverse)
library(xaringan)
library(PPBDS.data)
library(gganimate)
```

```{r, include = F}

gov <- qscores %>%
  filter(department == "GOV") 
  
comp_sci <- qscores %>%
  filter(department == "COMPSCI")

econ <- qscores %>%
  filter(department == "ECON")

## testing 10,000 times
n <- 10000 
t1err <- 0

## each test, make new 5-course sample from each department
for (i in 1:n){
   x <- sample(gov$hours, 5)
   y <- sample(econ$hours, 5)
   
## if p-value of t.test is less than 0.05, fail to reject the null and add 1 to number of Type 1 Errors
   if (((t.test(x,y, alternative = "two.sided"))$p.value)<=0.05) (t1err=t1err+1) 
}

t1_freq <- t1err / n

## Generate fake data for plotting
set.seed(12345)
fake_data_gov <- data.frame(x = rnorm(5, mean(gov$hours-1), 0.9), y = jitter(rep(0.05,5),amount = 0.05), dep = rep("GOV", 5))
fake_data_econ <- data.frame(x = rnorm(5, mean(econ$hours + 1), 0.9), y = jitter(rep(0.05,5),amount = 0.05), dep = rep("ECON", 5))
fake_data_1 <- fake_data_gov %>%
  full_join(fake_data_econ)

```

```{r p1, echo = F}

## fig1, blank plot

plot <- ggplot(fake_data_1,aes(x = x, y = y, colour = as.factor(dep))) +
  xlim(0,15) +  
  ylim(-0.05,0.3) +
  coord_cartesian(clip = "off") 

## fig2, add gov hours distribution

plot1 <- plot +
  stat_function(fun = dnorm, args = list(mean = mean(gov$hours), sd = sd(gov$hours)), colour = "#F8766D", size = 1.2) 

## fig3, add ec hours distribution

plot2 <- plot1 + 
  stat_function(fun = dnorm, args = list(mean = mean(econ$hours), sd = sd(econ$hours)), colour = "#7CAE00", size = 1.2) 

## fig4, jitter in fake data

plot3 <- plot2 +
  geom_jitter(size = 2, show.legend = F) +
  scale_color_manual(values = c("#F8766D", "#7CAE00"))

## fig5, boxplot for ranges

plot4 <- plot3 +
  geom_boxplot(data = fake_data_1, aes(x = x, y = y, colour = dep), varwidth = T, show.legend = F, orientation = "y")

```

<!-- align in center -->

.center[
# Type 1 Errors

```{r, echo = F}

## quick animation

## generate fake data, two identical groups, ten iterations
x <- data.frame(x = c(rnorm(50,6,3), rnorm(50,-6,3)), group = c(rep(1,50), rep(2,50)), y = rnorm(100,0,3), id = rep(seq(1:10),10))

## plot fake data, color by group

ggplot(x, aes(x = x, y = y, color = as.factor(group))) +
  ylim(-15,15) +
  xlim(-15,15) +
  geom_point(size = 2) +
  guides(color = F) +

  ## transition by iteration
  
  transition_states(id) +

## strip theme
  
  theme(axis.line=element_blank(),axis.text.x=element_blank(),
          axis.text.y=element_blank(),axis.ticks=element_blank(),
          axis.title.x=element_blank(),
          axis.title.y=element_blank()) +
  scale_color_manual(values = c("#F8766D", "#7CAE00"))

```
]
---

## We want to test if there is a statistically significant difference in means between the hours of courses in the Government department and courses in the Economics department. Looking at Q-guide data from Fall 2019, we see that Government courses average about 5.3 hours per week, and Economics courses average about 6 hours. This is not a very big difference, and a t-test on all this data would fail to reject the null. However, we can only see the reported hours for 5 randomly chosen courses in each department.


---

<!-- create a left column -->

.pull-left[

##### Assuming normality, let's plot the true distribution of course hours in the Government department, in red.
  

]

<!-- interval slide -->

--

<!-- create a right column -->

.pull-right[

```{r, echo = FALSE}
plot1
```

] 

---

.pull-left[

##### Now, in green, let's plot the distribution of courses in the Econ department.
]

.pull-right[
```{r, echo = FALSE}
plot1
```

] 

---


.pull-left[

##### Now, in green, let's plot the distribution of courses in the Econ department. 
]

.pull-right[
```{r, echo = FALSE}
plot2
```

] 
---


.pull-left[

##### There are many unique 5-course samples we could take from this data. One such sample could look like this:

]

.pull-right[
```{r, echo = FALSE}
plot2
```

]
---

<!-- create a left column -->

.pull-left[
##### There are many unique 5-course samples we could take from this data. One such sample could look like this:
]

.pull-right[
```{r, echo = FALSE}
plot3
```

] 
---
<!-- create a left column -->

.pull-left[
##### We see that our samples are really different, even though the full populations aren't. Using this sample, we'd conclude the true means were different and incorrectly reject the null hypothesis. This is a Type 1 error!
]

.pull-right[
```{r, echo = FALSE}
plot4
```

] 
---
<!-- create a left column -->

.pull-left[
##### If we simulate a real 5-course sampling of the government and economics departments, many times, we see that we will get a type 1 error, with statistical significance, about 2.5% of the time.
]

.pull-right[
```{r, echo = FALSE}
plot4
```

] 
---


