---
output_yaml:
  - _output.yml
---

# Continuous Response {#continuous-response}


<!-- Use nes with ideology as the lefthand side variable? (Should we just do 2016?) Models to use: lm(), stan_glm(), gam() -->

<!-- Overview: The main structure of the this chapter is three full scale linear modeling approaches to the same data set, where the response variable is continuous. (Maybe we have two data sets, and apply all three models to each one?) In each case, we use the tidymodels framework. I am not sure what the best dataset would be. I am not sure what the three approaches should be. Presumably lm() and stan_glm() would be included. Maybe gam() also. Or loess(). Or robust_lm(). Could be something fancier like svm(). Or neural network. -->


<!-- 0) Preamble. Plan out to chapter. Nothing more to learn. Just apply. Review themes.Rmd, and the three problems which we confronted in 9 and solved in 10. What specific questions are we going to answer? -->

<!-- 1) EDA of nes. drop na. Talk about Prudence at the end. It's own ### and with imgae in the side margin. OK if its only a paragraph or two.  -->

<!-- 2) Preview the three models. Justice. Review the Preceptor. Discuss Predictive. Might it be Causal. Show some math.  Courage, which is just code. -->

<!-- $$y_i = f(x_{i1}, x_{i2}, ..., \beta)$$ -->


<!-- 2) lm. Use all all variables. Go through the process. Use CV to provide a measure of how well it does. Use new_data. And make some predictions. Not noticing that this is a bad model because of overrfitting. -->

<!-- 3) stan_glm Do the same thing. Notice that CV measures are much better! Oh! Lesson: Be wary of overfitting. Maybe go back and re-estimate lm with fewer variables. Or not. -->

<!-- 4) Neural network. https://www.tidymodels.org/learn/models/parsnip-nnet/ All the same things apply. How well in a CV. predictions with new data.  -->

<!-- 5) How do we select among the models? Temperance -->

<!-- 6) Answer the questions we started with. -->


<!-- EC: Chapter Organization
Since Ch. 14 might be partially absorbed into this chapter, I'm wondering how to organize it.
Concepts in this chapter are multivariate regression, interaction models, parallel slope models, model selection, 3D scatterplots & regression planes. Concepts in Ch 14 are using tidymodels, machine learning, and cross validation. Here's a rough outline;
1. Multivariate regression and tidymodels
2. Interaction models, parallel slope models, model selection
3. 3D scatterplots & regression planes
4. Introducing machine learning
5. Using tidymodels for cross validation-->



<!-- 2. Use a different model, like loess, to solve the exact same problem. Go through the same overview, but more quickly. Not everything will work. For example, there are no parameter estimates for loess, or at least none that are easily visible. -->

<!-- 3. How do we decide? lm() or loess() or something? Incorporate (i.e., copy and paste) chapter 14 material. Don't do anything with tidymodel syntax. All that falls to chapter 11. But you are explaining every concept. -->

<!-- In loess section, grab a copy of this xkcd and use it: https://xkcd.com/2048/ -->

<!-- EC: Where is machine learning being introduced, if at all? -->

In the last two chapters, we covered regressions and common pitfalls. This chapter will be all about applying these concepts using the `tidymodels` framework.

The dataset we will be using is `nes` from the `PPBDS.data` package. `nes` contains data from the American National Election Survey, conducted every presidential election cycle. Along with demographic details, such as race, gender, and age, the survey also contains respondents' ideological identification. Because `ideology` is measured on a scale from 1 to 7, we can treat it as our continuous outcome variable.

Wouldn't it be interesting to predict `ideology` based off of other variables? This question has wide-reaching consequences for political polling and election outcomes. Throughout this chapter, we seek to answer this question: Which factors influences one's ideology, and how so?

## Exploratory Data Analysis

Load `PPBDS.data` and `skimr`. `skim()` the dataset.

```{r echo=FALSE, message=FALSE}
library(PPBDS.data)
library(tidyverse)
nes <- nes %>%
  mutate(state = as.factor(state),
         gender = as.factor(gender),
         race = as.factor(race),
         pres_appr = as.factor(pres_appr),
         education = as.character(education))
```

```{r}
library(PPBDS.data)
library(skimr)
skim(nes)
```

Great! Before we dive in, let's identify the variables of interest:

* `year`: the year the study was conducted
* `state`: the two-letter abbreviation corresponding to the state of the respondent
* `gender`: identifies respondents with values "Male" and "Female"
* `race`: race/ethnicity respondent identification
* `income`: integer variable representing 5 income groups: 1 as 0-16th percentile, 2 as 17-33rd, 3 as 34 to 67, 4 as 68 to 95, 5 as 96 to 100 
* `age`: ordered factor of respondents' age ranges
* `education`: ordered factor variable with a 7 tier delineation of educational achievement
* `pres_appr`: respondents' self-reported approval of the sitting president
* `ideology` as the continuous outcome variable, with 1 corresponding to strongly Democrat and 7 corresponding to strongly Republican and 0 if NA.

As you may have gleaned, these variables comprise a person's background. If we were given a new person who was not surveyed, and the above variables, our goal is to predict their ideology correctly. 

As you also saw, there are a lot of data points in this dataset. `nes` covers 1954 through 2016, but for our sake, we will narrow our scope to just 2016. Because of the changing nature of what makes someone ideologically liberal or ideologically conservative over time, it is best to construct a model in which we eliminate as much of this variance as possible. Keep in mind that, when feeding in new data to our model, it is most accurately applied to voters in or around 2016. Indeed, a 1954 voter might act very differently than a 2016 voter.

Filter the data to only show `year == 2016`.

```{r}
library(tidyverse)
nes %>%
  select(year, state, gender, race, income, age, education, ideology) %>%
  filter(year == 2016)
```

Note from our previous skimming of the data that all of the variables are incomplete, meaning they contain `NA` values. While there are methods to impute missing data, we will simply remove these values for now. We will save this filtered and cleaned dataset as `nes_2016`.

```{r}
nes_2016 <- nes %>%
  filter(year == 2016) %>%
  select(-year) %>%
  drop_na()

nes_2016
```

With this cleaned data, let's dive into the first model we will create, which will create a linear regression.

## `lm()`

First, load the ***tidymodels*** library and use `initial_split()` to split the data into testing and training sets. Don't worry about the testing set-- for now,  we will build our model off of the training set.

```{r eval=FALSE}
library(tidymodels)
library(rsample)

set.seed(10)
nes_split <- initial_split(nes_2016)

nes_training <- nes_split %>% training()
nes_testing <- nes_split %>% testing()
```

```{r}
library(tidymodels)
library(rsample)

set.seed(10)
nes_split <- initial_split(nes_2016)

nes_training <- nes_split %>%
  training() %>%
  mutate(education = as.character(education))
nes_testing <- nes_split %>%
  testing() %>%
  mutate(education = as.character(education))
```

Next, construct the linear regression model.

```{r}
lm_model <-
    linear_reg() %>%
    set_engine("lm") %>%
    set_mode("regression")
```

Here, we are using the ***parsnip*** package to create an engine that can run this linear regression model easily and repeatedly. `linear_reg()` tells the engine that this is a linear regression. `set_engine("lm")` tells the engine to use the `lm()` function. `set_mode()` has to options, "regression" and "classification". Since our left-hand variable is a continuous one, we will set it to "regression". If it were categorical, we would set the mode to "classification".

Fit the model to the data. We will first create a regression formula and save it as an R object for future recall. There are a lot of different combinations we could use for the formula, but which will yield the *best* model? For now, we will build a full model containing all of the predictor variables

```{r}
interact_form <- formula(ideology ~ gender * race + state * income + age + education + pres_appr)
```

```{r eval=FALSE}
fit(lm_model, interact_form, nes_training)
```

Note that we can use `tidy()`, just like we did in previous chapters, to take a look at the results:

```{r}
fit(lm_model, interact_form, nes_training) %>%
  tidy(conf.int = TRUE) %>%
  select(term, estimate, conf.low, conf.high)
```

How do we know if this is a good model? We can check using cross-validation. Cross-validation seeks to minimizes the MSE (mean squared error) of its predicted and real outcomes. There are two important characteristics of the MSE we should always keep in mind:

1. We can think our estimate of the MSE is a random variable. For example, the dataset we have may be a random sample from a larger population. An algorithm may have a lower apparent error than another algorithm due to luck.

2. If we train an algorithm on the same dataset that we use to compute the MSE, we might be overtraining. In general, when we do this, the apparent error will be an underestimate of the true error.

Cross validation is a technique that permits us to alleviate both these problems. To understand cross validation, it helps to think of the *true error*, a theoretical quantity, as the average of many *apparent errors* obtained by applying the algorithm to new random samples of the data, none of them used to train the algorithm. 

However, we only have available one set of outcomes: the ones we actually observed. Cross validation is based on the idea of generating a series of different random samples on which to apply our algorithm. There are several approaches we can use, but the general idea for all of them is to randomly generate smaller datasets that are not used for training, and instead used to estimate the true error.

### K-Fold Cross-validation

Generally speaking, a machine learning challenge starts with a dataset. We need to build an algorithm using this dataset that will eventually be used in completely independent datasets.

So to imitate this situation, we carve out a piece of our dataset and pretend it is an independent dataset: previously, we divided the dataset into a _training set_ and a _test set_. We will train our algorithm exclusively on the training set and use the test set *only for evaluation purposes*.

We usually try to select a small piece of the dataset so that we have as much data as possible to train. However, we also want the test set to be large so that we obtain a stable estimate of the loss without fitting an impractical number of models. The `initial_split()` function reserves 25% of the data for testing by default. 

Remember, we cannot touch the testing set! One way we can check whether the model we created works is to use cross-validation, which avoids the problem of overtraining by splitting the data into smaller sections. We'll show you how to do using the ***rsample*** package.

The function `vfold_cv()` splits your training set into `v`, a specified number of smaller sections. The most common values for `v` are 5 and 10. To keep things simple, we will be using `v = 5` to keep the individual sections large enough to accurately train the models. The first four folds will be used to train the model, and the last fold is always used to assess the accuracy of the model.

Below, we will create a new object called `nes_folds` that contains equally sized sections of `nes_training`.

```{r}
set.seed(10)
nes_folds <- nes_training %>%
  vfold_cv(v = 5)
```

How can we work with the `nes_folds` object?  **tidymodels** makes it easy by using the `fit_resamples()` function in the **tune** package. The `fit_resamples()` function takes as its first argument a model specification (such as `lm_model`). It takes a formula as its second argument, either in traditional form or as a `recipe()`. 

```{marginfigure, echo=TRUE}
Recipes are what ***tidymodels*** use in lieu of defining a formula traditionally. `recipe()` is useful because you can define both the formula and any transformations you want to make with the data. Recipes are easily called and re-used, making it easy to use the same recipe with different models.

There are a number of useful functions that can be piped into recipes. One particularly useful family of functions is the family of step functions. These functions conduct a variety of transformations, from removing variables with high correlations to other variables (`step_corr()`) to adding in an interaction term (`step_interact()`). To learn more about recipes, you can visit the [Tidymodels website.](https://recipes.tidymodels.org/reference/index.html)
```

Usually, a simple formula without interaction terms can be plugged right in to `fit_resamples()`. However, ***tidymodels*** does not allow in-line functions in recipes, meaning we need to use a `step_*` function to add the interaction terms. Let's create our recipe.

```{r}
lm_rec <- recipe(ideology ~ gender + race + state + income + age + education + pres_appr, data = nes_training) %>%
  step_interact(terms = ~ gender:race + state:income)
```

<!-- EC: Adding interaction terms seems to complicate fit_resamples. -->

Finally, the `resamples` argument is where you input the cross-validation dataset.

<!-- EC: Not sure why fold 2 is getting an error but the other folds are not. -->

```{r}
set.seed(10)
fit_resamples(object = lm_model,
              preprocessor = lm_rec,
              resamples = nes_folds)
```   

Because the metrics are in list-columns, we can extract the average metrics across all the folds using the `collect_metrics()` function:

```{r}
set.seed(10)
fit_resamples(lm_model,
              lm_rec,
              nes_folds) %>%
  collect_metrics()
```

Now that we've viewed the cross-validation metrics, it's time to use new data: namely, the testing set. Remember, we only touch the testing set for evaluation purposes. To apply our model to `nes_testing`, use the `predict()` function.

```{r}
lm_model %>%
  fit(interact_form, data = nes_training) %>%
  predict(new_data = nes_testing)
```

To extract the rmse, we set the "truth" to `ideology` so this function can compare our predicted values to the true values.

```{r}
lm_model %>%
  fit(interact_form, data = nes_training) %>%
  predict(new_data = nes_testing) %>%
  bind_cols(nes_testing) %>%
  rmse(truth = ideology, estimate = .pred)
```

The mean squared error is about 1.64.

What if we create completely new data outside of the testing set? Time to make up some imaginary people. Let's say we have four individuals whose ideology in 2016 we wanted to predict. We can create a tibble with the values of their demographic information, like so:

<!-- EC: Do we want to include pres_appr? Is that a meaningful variable or is that too close to the outcome variable? -->

```{r}
new_people <- tibble("Name" = c("Alice", "Bob", "Chelsea", "Darren"),
                     "state" = as.factor(c("VA", "AL", "VA", "AL")),
                     "gender" = as.factor(c("Female", "Male", "Female", "Male")),
                     "income" = c(3, 1, 3, 5),
                     "age" = c("35 - 44", "55 - 64", "35 - 44", "55 - 64"),
                     "education" = c("elementary", "hs diploma", "hs diploma +", "hs diploma"),
                     "race" = c("White", "White", "White", "White"),
                     "pres_appr" = as.factor(c("approve", "approve", "approve", "approve")))
```

Now, let's predict each new person's ideology using the linear regression model we just created.

```{r this-is-not-knitting, eval=FALSE}
lm_model %>%
  fit(interact_form, data = nes_training) %>%
  predict(new_data = new_people) %>%
  bind_cols(new_people) %>%
  rename("ideology" = ".pred")
```

As you can see, the two women, Alice and Chelsea, have similar values for their variables save for one: education. Alice is educated at the elementary level, whereas Chelsea holds an advanced degree. Their predicted ideologies are 2.22 and 2.42 respectively, showing that a 35-44 year old white woman from Virginia who approved of the sitting President in 2016 with an advanced degree was on average 0.20 more conservative than her counterpart with an elementary degree.

Meanwhile, white 55-64 year old men with high school diplomas from Alabama differ in their income; Bob is two standard deviations above average, whereas Darren is two standard deviations below average in earnings. All else held equal, Bob's ideology was estimated to be more conservative than Darren's by 0.13. Let's see if these observations hold up with our other models.

## Using `stan_glm()`

The ***rstanarm*** package contains a lot of powerful functions that conduct Bayesian data analysis by using priors. One such function is `stan_glm()`, which you can think of as the Bayesian way of fitting a regression model. 

We will be following the exact same steps as before. First, we will construct a ***parsnip*** model that uses the "stan" engine.

```{r message=FALSE}
library(rstanarm)

stan_model <-
    linear_reg() %>%
    set_engine("stan") %>%
    set_mode("regression")
```

Next, let's fit this model using the formula we specified before. We can select for the relevant terms after using `tidy()` to make our tibble viewable.

```{r message=FALSE, warning=FALSE, eval=FALSE}
set.seed(10)
fit(stan_model, interact_form, nes_training) %>%
  tidy(conf.int = TRUE) %>%
  select(term, estimate, std.error)
```

<!-- EC: Removed LOO. Add back in K-fold -->

### Verifying fit using cross-validation

Let's practice K-Fold validation with the `stan` model.

```{r cache=TRUE}
set.seed(10)
fit_resamples(object = stan_model,
              preprocessor = lm_rec,
              resamples = nes_folds)
```   


```{r cache=TRUE, eval=FALSE}
set.seed(10)
fit_resamples(stan_model,
              lm_rec,
              nes_folds) %>%
  collect_metrics()
```

Let's test the stan model using our four `new_people`.

```{r message=FALSE, eval=FALSE}
stan_model %>%
  fit(interact_form, data = nes_training) %>%
  predict(new_people) %>%
  bind_cols(new_people) %>%
  rename("ideology" = ".pred")
```

What if we created a bunch of different formulas that used different permutations of predictor variables? For example, one would exclude `state`, which has 50 different levels.

First, let's create a basic formula that only takes in `pres_appr`.

```{r}
basic_form <- formula(ideology ~ pres_appr)
```

Next, we can use `update()` to create the more complicated formulas.  `update()` takes as its first argument a formula and as its second argument the additions you want to make.  To keep all the predictors from the first formula and add more, you will start with `~ . + ` and then add more predictors, like so:

```{r}
race_gender_form <- update(basic_form,
                    ~ . + race + gender)

demo_form <- formula(ideology ~ age + gender + race + income)

non_state_form <- update(demo_form,
                         ~ . + education + pres_appr)

full_form <- update(non_state_form,
                   ~ . + state)
```

Next, we will save these different formulas, along with the original `interact_form`, in a tibble to easily remember and recall the formulas.

```{r}
nes_formulas <- tibble(formula = c(basic_form,
                                   race_gender_form,
                                   demo_form,
                                   non_state_form,
                                   full_form,
                                   interact_form),
                       group = c("Presidential approval model",
                                 "Race and gender model",
                                 "Demographic model",
                                 "Full model without state",
                                 "Full model",
                                 "Interaction model"))
```

Now, we can use `map_*` to apply all of these models and view their metrics to see which ones have the lowest rmse values.

```{r cache=TRUE}
set.seed(10)
folds_metrics<- nes_formulas %>%
  mutate(metrics = map(formula, ~ fit_resamples(object = lm_model,
                                                preprocessor = .,
                                                resamples = nes_folds) %>%
                         collect_metrics()))
```

Let's present the results stored in our `folds_metrics` object:

```{r message=FALSE, warning=FALSE}
folds_metrics %>%
  mutate(mean_rmse = map_dbl(metrics, ~ filter(., .metric == "rmse") %>% pull(mean)),
         se_rmse = map_dbl(metrics, ~ filter(., .metric == "rmse") %>% pull(std_err))) %>%
  select(group, mean_rmse, se_rmse)
```

Looking at the results, it seems like `pres_appr` is a good predictor variable to incorporate, as the rmse for `basic_form` is much lower than that of the Demographic model, which does not include `pres_appr`. This makes sense, as how you lean ideologically is very likely correlated with how you feel about the President.

The models with the lowest rmse values seem to be the Full model excluding state, which includes:

* `pres_appr`
* `education`
* `race`
* `gender`
* `age`
* `income`

It is very similar in rmse value with the Education model, which excludes `age` and `income`.

Let's test the Full model excluding state on the testing set.

```{r}
lm_model %>%
  fit(non_state_form, data = nes_training) %>%
  predict(new_data = nes_testing)
```

To extract the rmse, we set the "truth" to `ideology` so this function can compare our predicted values to the true values.

```{r}
lm_model %>%
  fit(non_state_form, data = nes_training) %>%
  predict(new_data = nes_testing) %>%
  bind_cols(nes_testing) %>%
  rmse(truth = ideology, estimate = .pred)
```

Discuss rmse value

## Using a neural network

<!-- EC: Is this particular approach better suited for classification questions? (chapter 12) https://www.tidymodels.org/learn/models/parsnip-nnet/ -->

Let's dive into the world of machine learning. Neural networks are useful for mimicking how humans make decisions. Thanks to the ***parsnip*** package, there are functions available that can fit regression models using neural networks.

Let's create the neural network engine.

```{r }
nnet_model <- mlp(hidden_units = 5) %>%
  set_mode("regression") %>%
  set_engine("nnet") %>%
  translate()
nnet_model
```

### Verifying fit with cross-validation

Let's cross-validate our predictions using `vfold_cv()`.

```{r message=FALSE, eval=FALSE}
set.seed(10)
fit_resamples(object = nnet_model,
              preprocessor = lm_rec,
              resamples = nes_folds)
```   

```{r eval=FALSE}
set.seed(10)
fit_resamples(nnet_model,
              lm_rec,
              nes_folds) %>%
  collect_metrics()
```

### Predicting new data

Lastly, let's apply our neural network model to the testing set.

```{r}
nnet_model %>%
  fit(non_state_form, data = nes_training) %>%
  predict(new_data = nes_testing) %>%
  bind_cols(nes_testing) %>%
  rmse(truth = ideology, estimate = .pred)
```

Now, let's predict each new person's ideology using the `nnet` model we created, like before.

```{r message=FALSE, eval=FALSE}
nnet_model %>%
  fit(interact_form, data = nes_training) %>%
  predict(new_people) %>%
  bind_cols(new_people) %>%
  rename("ideology" = ".pred")
```

## Model selection

By now, we have created three different linear regression models: one using the `lm` engine, one using the Bayesian `stan` engine, and one using the machine learning `nnet` engine. We have also explored three different methods to analyze the accuracy of our models' out-of-sample predictions: K-fold cross-validation, Leave-One-Out cross-validation, and bootstrapping. We've also tested six different formulas.

So many options! How do we choose which model and formula is the best?

First, we should only favor more complex models (or formulas) if the additional complexity is *warranted*. This is a philosophical principle known as "Occam's Razor." It states that, "all other things being equal, simpler solutions are more likely to be correct than complex ones." When viewed in a modeling framework, Occam's Razor can be restated as, "all other things being equal, simpler models are to be preferred over complex ones." In other words, we should only favor the more complex model/formula if the additional complexity is *warranted*.

Secondly, we have two quantitative metrics to help guide our decision: the `rmse` and `rsq` values. Lower `rmse` values are more desired because they indicate a higher accuracy in predicting the outcome variable. Higher `rsq` values are better because that means that more of the variation in the dataset can be explained by the predictor variables in the formula.


## Wrap-Up

So, how can we predict someone's ideology based off of other information?

* `pres_appr`
* trends in `age`, `income`, `gender`, `race`, `education`
* caveat interaction variables
* caveat applying to other data before 2016
