---
output_yaml:
  - _output.yml
---


# Getting Help {-}


<!-- Maybe useful: https://github.com/mine-cetinkaya-rundel/rmd-reprex -->

*The best data science superpower is knowing how to ask a question.* -- Mara Averick

## Searching for Help with R

Google is your friend. Searching for help, especially when you are not sure what you are looking for, is more art than science. Adding certain strings to the search query --- like "R", "tidyverse", or "ggplot" --- often helps. Some locations are high quality (anything to do with RStudio or tidyverse) while others are less good than they initially appear (sthda.com, r-statistics.co, rdocumentation.org).

## Reproducible Examples

The best way to get help from other people is to create a **rep**roducible **ex**ample --- a so-called "reprex" --- which shows an example of the problem you are having. 

Install the **reprex** package:

```{r, eval=FALSE}
install.packages("reprex")
```

Load the library:

```{r}
library(reprex)
```

The code for your reproducible example needs to exist somewhere *before* you can use it to create your reprex. It is tempting to just grab it from the current file in which you are working, usually either a .R or .Rmd. Instead, I recommend creating a new file, dedicated to the example. Call it whatever.R. This makes everything easier, not least because some of the code you need, like the calls to *library* will often be many lines away, in your original file, from the code which is generating your error. Creating a stand-alone file helps to make the problem clear in your own head. Example:

```{r, eval=FALSE}
library(tidyverse)
mtcars %>% 
  mutate(new_var = mpg + cly)
```

Running this code produces an error because I have misspelled `cyl` as `cly`. Try it! To create a reprex:

1. Highlight the code.
1. Press Ctrl-C (or the equivalent) to copy the code to the clipboard.
1. Type `reprex()` at the R prompt.

The reproducible example now exists in two places. First, you can see it in the Viewer pane in RStudio. Second, it is on your clipboard. 

Next step is to go to [RStudio Community](https://community.rstudio.com/) (or whatever location at which you want to ask your question), create a new post, and paste in the reprex from your clipboard. Of course, you may first have to create an account. Also, your post will need a title, associated text and perhaps other information like tags. The key point is that using a reproducible example makes it much likelier that someone will answer your question.

[Here](https://community.rstudio.com/t/best-practices-how-to-prepare-your-own-data-for-use-in-a-reprex-if-you-can-t-or-don-t-know-how-to-reproduce-a-problem-with-a-built-in-dataset/5346) is advice about using your own data in making a reprex. [Datapasta](https://cran.r-project.org/web/packages/datapasta/vignettes/how-to-datapasta.html) may also be relevant. 
