---
output_yaml:
  - _output.yml
---

# M and S Errors {-}

<!-- Somewhat concerned about errors caused by the failure to include Tufte fonts, as in Robert's previous header. -->

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tufte)
```

To navigate tutorials, click on the box below and use your left and right arrow keys to transition between slides. 

## Traditional Errors

Under the null hypothesis significance testing framework, or 'NHST', researchers have to be cautious about two types of error. A *Type 1* error, or "false positive" is the rejection of a true null hypothesis, and a *Type 2* error, or "false negative" is a failure to reject the null despite the alternate hypothesis being true. 


### Type 1 Errors

<br/>
```{r embed_flip1, echo = FALSE, fig.align='left', fig.cap="Type 1 errors occur when our samples look different, even though the underlying distributions from which they are drawn are similar."}

## important to align left so margin comments can fit

## embed the flipbook linked from 

knitr::include_url('m-s-errors/type1_flip.html')

```

<br/><br/>

### Type 2 Errors

<br/>
```{r embed_flip2, echo = FALSE, fig.align='left', fig.cap = "Type 2 errors occur when our samples look similar, even though the underlying distributions from which they are drawn are different."}



# embed the flipbook

knitr::include_url('m-s-errors/type_2_flip.html')

```

<br/><br/>

## Dangers of the NHST Framework

Recent work has shown that this model of testing errors fails to capture all the risks of hypothesis testing. A new method, first proposed by Gelman and Carlin (2014) establishes two new types of errors that they think will lead to more robust and reproducible results. They think that researchers should attempt to measure the probability they will get the sign wrong on their estimate, a *Type S* error, and the factor by which their estimate overestimates the magnitude of the true effect, a *Type M* error. Under the framework of statistical significance, studies will be very prone to their new errors.

### Type M and Type S Errors

<br/>
```{r embed_sm, echo = FALSE, fig.align='left'}


knitr::include_url('m-s-errors/sm_flip.html')

```

<br/><br/>

### How prevalent are Type M and Type S errors? Why are they important?

<br/>
```{r embed_power, echo = FALSE, fig.align='left'}

knitr::include_url('m-s-errors/power_flip.html')

```

