# (APPENDIX) Appendix {-}

# Rubin Causal Model

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(knitr)
library(kableExtra)
library(tidyverse)
```

The Rubin causal model (RCM) is an approach to the statistical analysis of cause and effect based on the framework of potential outcomes, named after Donald Rubin.^[This chapter remixes some material from [Rubin causal model (Wikipedia)](https://en.wikipedia.org/wiki/Rubin_causal_model). <br /> The name "Rubin causal model" was first coined by Paul W. Holland (1986).  The potential outcomes framework was first proposed by Jerzy Neyman in his 1923 Master's thesis (Neyman 1923), though he discussed it only in the context of completely randomized experiments (Rubin 2005)..]  

<!-- Perhaps the credit to Wikipedia should be added to the attributions page
rather than here. -->

<!-- Rewrote this to stick with one example (headache) -->

The Rubin causal model (RCM) is based on the idea of **potential outcomes.**  For example, let's say you took aspirin and now no longer have a headache. To measure the causal effect of taking the aspirin, we need to compare the outcome for you in one alternative future (where you took the aspirin) to another (where you did not). Since it is impossible to see both potential outcomes at once, one of the potential outcomes is always missing. This dilemma is the "fundamental problem of causal inference."

## Causal effects

Here's how Rubin defines a "causal effect": 

> Intuitively, the causal effect of one treatment, E, over another, C, for a particular unit and an interval of time from $t_1$ to $t_2$ is the difference between what would have happened at time $t_2$ if the unit had been exposed to E initiated at $t_1$ and what would have happened at $t_2$ if the unit had been exposed to C initiated at $t_1$: 'If an hour ago I had taken two aspirins instead of just a glass of water, my headache would now be gone,' or 'because an hour ago I took two aspirins instead of just a glass of water, my headache is now gone.' Our definition of the causal effect of the E versus C treatment will reflect this intuitive meaning.^[Rubin 1974.]

Thus, according to the RCM, the causal effect of your taking or not taking aspirin one hour ago is the difference between how your head would have felt in case 1 (taking the aspirin) and case 2 (not taking the aspirin). If your headache would remain without aspirin but disappear if you took aspirin, then the causal effect of taking aspirin is headache relief.

In most circumstances, we are interested in comparing two futures, one generally termed "treatment" and the other "control." These labels are somewhat arbitrary. The difference between the potential outcome under treatment and the potential outcome under control is called a "causal effect" or a "treatment effect."  The scenario that didn't actually happen, and thus that we didn't observe, is called a "counterfactual."

## Potential outcomes

Suppose that Joe is participating in an FDA test for a new hypertension drug. If we were omniscient, we would know the outcomes for Joe under both treatment (the new drug) and control (either no treatment or the current standard treatment). The causal effect, or treatment effect, is the difference between these two potential outcomes. 

<!-- Should we stick with this notation? -->

```{r first_example, echo = FALSE}
# First, we create a tibble with the values we want for the table

tibble(`Subject` = "Joe",
       `$Y_t(u)$` = c("$130$"),
       `$Y_c(u)$` = c("$135$"),
       `$Y_t(u) - Y_c(u)$` = c("$-5$")) %>%
  
  # Then, we use the kable function to make it pretty
  
  kable() %>%
  kable_styling(bootstrap_options = "striped",
                full_width = FALSE,
                position = "left") %>%
  column_spec(1,
              bold = TRUE,
              border_right = TRUE)
```

<!-- Should the numbers be written in math font? -->

$Y_{t}(u)$ is Joe's systolic blood pressure if he takes the new pill. In general, this notation expresses the potential outcome which results from a treatment, $t$, on a unit, $u$. Similarly, $Y_{c}(u)$ is the effect of a different treatment, $c$ or control, on a unit, $u$. In this case, $Y_{c}(u)$ is Joe's blood pressure if he doesn't take the pill. $Y_{t}(u)-Y_{c}(u)$ is the causal effect of taking the new drug.  Thus, Joe would have a blood pressure of $130$ if he took the pill and $135$ if he did not; the causal effect of taking the pill is $-5$.  Joe should take the pill!

From this table we only know the causal effect on Joe. Everyone else in the study might have an increase in blood pressure if they take the pill. However, regardless of what the causal effect is for the other subjects, the causal effect for Joe is lower blood pressure, relative to what his blood pressure would have been if he had not taken the pill.

## No causation without manipulation

In order for a potential outcome to make sense, it must be possible, at least *a priori*. For example, if there is no way for Joe, under any circumstance, to obtain the new drug, then $Y_{t}(u)$ is impossible for him. It can never happen. And if $Y_{t}(u)$ can never be observed, even in theory, then the causal effect of treatment on Joe's blood pressure is not defined. 

The causal effect of new drug is well defined because it is the simple difference of two potential outcomes, both of which might happen. In this case, we (or something else) can manipulate the world, at least conceptually, so that it is possible that one thing or a different thing might happen.

This definition of causal effects becomes much more problematic if there is no way for one of the potential outcomes to happen, ever. For example, what is the causal effect of Joe's height on his weight? Naively, this seems similar to our other examples. We just need to compare two potential outcomes: what would Joe's weight be under the treatment (where treatment is defined as being 3 inches taller) and what would Joe's weight be under the control (where control is defined as his current height).

A moment's reflection highlights the problem: we can't increase Joe's height. There is no way to observe, even conceptually, what Joe's weight would be if he were taller because there is no way to make him taller. We can't manipulate Joe's height, so it makes no sense to investigate the causal effect of height on weight. Hence the slogan: **No causation without manipulation.**

## Average treatment effect

<!-- Calling this the "average treatment effect" rather than "average causal
effect" throughout -->

Consider a larger sample of patients: 

```{r average_treatment_effect, echo = FALSE}
# First, we create a tibble with the values we want for the table

tibble(`Subject` = c("Joe", "Mary", "Sally", "Bob", "James", "MEAN"),
       `$Y_t(u)$` = c("$130$", "$130$", "$130$", "$140$", "$145$", "$135$"),
       `$Y_c(u)$` = c("$135$", "$145$", "$145$", "$150$", "$140$", "$143$"),
       `$Y_t(u) - Y_c(u)$` = c("$-5$", "$-15$", "$-15$", "$-10$", "$+5$", "$-8$")) %>%
  
  # Then, we use the kable function to make it pretty
  
  kable() %>%
  kable_styling(bootstrap_options = "striped",
                full_width = FALSE,
                position = "left") %>%
  column_spec(1,
              bold = TRUE,
              border_right = TRUE)
```

The causal effect is different for every subject, but the drug works for Joe, Mary, Sally and Bob because the causal effect is negative. Their blood pressure is lower with the drug than it would have been if each did not take the drug. For James, however, the drug causes an *increase* in blood pressure. 

One may calculate the **average treatment effect** (often abbreviated ATE) by taking the mean of all the causal effects.  Here, the mean is $-8$.

Note that we can calculate the average treatment effect in this case because we are omniscient and know all the potential outcomes under both treatment and control.  We will soon see what assumptions we need to make in order to calculate the average treatment effect in the real world where only one potential outcome can ever be observed.  In particular, while we cannot estimate individual causal effects without very strong assumptions, randomized experiments can allow for good estimates of the ATE, as we will see later in this chapter.

## Stable unit treatment value assumption (SUTVA)

One important assumption for causal inference is that "the [potential outcome] observation on one unit should be unaffected by the particular assignment of treatments to the other units."^[Cox 1958, §2.4.] This is called the **Stable Unit Treatment Value Assumption** (**SUTVA**). 

In the context of our example, Joe's blood pressure should not depend on whether Mary receives the drug. But what if it does? Suppose that Joe and Mary live in the same house and Mary always cooks. The drug causes Mary to crave salty foods, so if she takes the drug she will cook with more salt than she would have otherwise. A high salt diet increases Joe's blood pressure. Therefore, his outcome will depend on both which treatment he received and which treatment Mary receives.

SUTVA violation makes causal inference more difficult. We can account for dependent observations by considering more treatments. We create 4 treatments by taking into account whether or not Mary receives treatment:

```{r joe_mary, echo = FALSE}
# First, we create a tibble with the values we want for the table

tibble(`Subject` = c("Joe"),
       `Joe = c,<br />Mary = t` = c("$140$"),
       `Joe = t,<br />Mary = t` = c("$130$"),
       `Joe = c,<br />Mary = c` = c("$125$"),
       `Joe = t,<br />Mary = c` = c("$120$")) %>%
  
  # Then, we use the kable function to make it pretty. escape = FALSE is also
  # required for the linebreaks
  
  kable(escape = FALSE) %>%
  
  kable_styling(bootstrap_options = "striped",
                full_width = FALSE,
                position = "left") %>%
  column_spec(1,
              bold = TRUE,
              border_right = TRUE)
```

Recall that a causal effect is defined as the difference between two potential outcomes. In this case, there are multiple causal effects because there are more than two potential outcomes:

- One is the causal effect of the drug on Joe when Mary receives treatment ($130-140$).
- Another is the causal effect of the drug on Joe when Mary does not receive treatment ($120-125$).
- A third is the causal effect of Mary's treatment on Joe when Joe is not treated ($140-125$). In other words, we can define a causal effect on Joe even in situation in which Joe's treatment is identical in both situations.

Note that Mary taking the drug has a larger causal effect on Joe than Joe himself taking the drug, and what's more, the effect is in the opposite dierction!

By considering more potential outcomes in this way, we can cause SUTVA to hold. However, if any units other than Joe are dependent on Mary, then we must consider further potential outcomes. The greater the number of dependent units, the more potential outcomes we must consider and the more complex the calculations become (consider an experiment with 20 different people, each of whose treatment status can affect outcomes for every one else). In order (easily) to estimate the causal effect of a single treatment relative to a control, SUTVA should hold. 

## The fundamental problem of causal inference

The results we have seen up to this point could never be measured in practice. It is impossible, by definition, to observe the effects of more than one treatment on a single subject. Joe cannot both take the pill and not take the pill at the same time! Therefore, the data we actually observe would look something like this: 

```{r random_assignment, echo = FALSE}
# First, we create a tibble with the values we want for the table

tibble(`Subject` = c("Joe", "Mary", "Sally", "Bob", "James", "MEAN"),
       `$Y_t(u)$` = c("$130$", "$120$", "$?$", "$?$", "$115$", "$121.67$"),
       `$Y_c(u)$` = c("$?$", "$?$", "$125$", "$130$", "$?$", "$127.5$"),
       `$Y_t(u) - Y_c(u)$` = c("$?$", "$?$", "$?$", "$?$", "$?$", "$-5.83$")) %>%
  
  # Then, we use the kable function to make it pretty
  
  kable() %>%
  kable_styling(bootstrap_options = "striped",
                full_width = FALSE,
                position = "left") %>%
  column_spec(1,
              bold = TRUE,
              border_right = TRUE)
```

Question marks are responses that could not be observed. The **Fundamental Problem of Causal Inference**^[Holland 1986.] is that directly observing unit-level causal effects is impossible. However, this does not make *causal inference* impossible. Certain techniques and assumptions allow the fundamental problem to be overcome.

In particular, we may want to know whether we can get a good estimate of the average treatment effect.  If we knew that the drug lowered blood pressure on average, that would help us make decisions even if we can't observe the causal effect of the drug on individual people (because we don't observe counterfactuals).

The most natural place to start is with the difference in the sample mean outcomes between treatment and control, $-5.83$.  Note that unlike earlier, this is an *estimate* of the ATE, not the ATE itself.  Why?  Because we have missing data!  For each unit, we only observe one of the potential outcomes.

## The assignment mechanism

Is the difference in sample means between treated units and control units a good estimate of the average treatment effect?  That depends entirely on the method by which units are assigned treatment, which is called the **assignment mechanism**.

One such assignment mechanism is **randomization**. For each subject we could flip a coin to determine if she receives treatment. If we wanted five subjects to receive treatment, we could assign treatment to the first five names we pick out of a hat.

Randomized assignment is the best assignment mechanism for inferring average treatment effects because if the sample is large enough, the difference in sample means between treated and control units will be very close to the true average treatment effect in that sample.

However, when the sample size is small, our estimated average treatment effect may deviate considerably from the truth.  For example, the data in the last section were from just one possible random assignment, where Joe, Mary and James receive the treatment.  Here's another possible random assignment, where Joe, Mary and Sally receive the treatment:

```{r random_assignment_2, echo = FALSE}
# First, we create a tibble with the values we want for the table

tibble(`Subject` = c("Joe", "Mary", "Sally", "Bob", "James", "MEAN"),
       `$Y_t(u)$` = c("$130$", "$120$", "$100$", "$?$", "$?$", "$115$"),
       `$Y_c(u)$` = c("$?$", "$?$", "$?$", "$130$", "$120$", "$123$"),
       `$Y_t(u) - Y_c(u)$` = c("$?$", "$?$", "$?$", "$?$", "$?$", "$-8.33$")) %>%
  
  # Then, we use the kable function to make it pretty
  
  kable() %>%
  kable_styling(bootstrap_options = "striped",
                full_width = FALSE,
                position = "left") %>%
  column_spec(1,
              bold = TRUE,
              border_right = TRUE)
```

Note that with one assignment, the estimate of the ATE is $-5.83$, while with another it is $-8.33$.  Assume that these data are the truth: 

```{r true_data, echo = FALSE}
# First, we create a tibble with the values we want for the table

tibble(`Subject` = c("Joe", "Mary", "Sally", "Bob", "James", "MEAN"),
       `$Y_t(u)$` = c("$130$", "$120$", "$100$", "$110$", "$115$", "$115$"),
       `$Y_c(u)$` = c("$115$", "$125$", "$125$", "$130$", "$120$", "$123$"),
       `$Y_t(u) - Y_c(u)$` = c("$+15$", "$-5$", "$-25$", "$-20$", "$-5$", "$-8$")) %>%
  
  # Then, we use the kable function to make it pretty
  
  kable() %>%
  kable_styling(bootstrap_options = "striped",
                full_width = FALSE,
                position = "left") %>%
  column_spec(1,
              bold = TRUE,
              border_right = TRUE)
```

Thus, the true average treatment effect is $-8$.  However, our estimates of the average treatment effect vary because our sample is small and the responses have a large variance. If the sample were larger and the variance were less, the average treatment effect would be closer to the true average treatment effect regardless of the specific units randomly assigned to treatment.

## Permutation tests

<!-- This section attempts to describe permutation tests in a non-frequentist
way; thus, there are no references to a "test statistic" or to the "sharp null."
-->

Let's say we observed the following results, which we know came from a random treatment assignment:

```{r permutation_setup, echo = FALSE}
# First, we create a tibble with the values we want for the table

tibble(`Subject` = c("Mary", "Sally", "Bob", "James", "MEAN"),
       `$Y_t(u)$` = c("$120$", "$100$", "$?$", "$?$", "$110$"),
       `$Y_c(u)$` = c("$?$", "$?$", "$130$", "$120$", "$125$"),
       `$Y_t(u) - Y_c(u)$` = c("$?$", "$?$", "$?$", "$?$", "$-15$")) %>%
  
  # Then, we use the kable function to make it pretty
  
  kable() %>%
  kable_styling(bootstrap_options = "striped",
                full_width = FALSE,
                position = "left") %>%
  column_spec(1,
              bold = TRUE,
              border_right = TRUE)
```

The drug looks very useful!  However, how likely is it that we would observe these results if the drug did nothing at all?  That is, how confident should we be that the drug is actually helpful?

To answer this question, we can use a **permutation test**.^[See [this post](https://www.r-bloggers.com/what-is-a-permutation-test/) for a longer discussion.]  The intuition behind a permutation test is simple.  We observed four outcomes, two of which were assigned to treatment and two that were assigned to control.  To conduct a permutation test, we calculate our quantity of interest (here, the difference in means between treated and control) for every possible arrangement of the labels "treatment" and "control" across these four numbers.

This is easiest to understand visually:

```{r all_permutations, echo = FALSE}
# First, we create a tibble with the values we want for the table

tibble(`Permutation` = c("#1", "#2", "#3", "#4", "#5", "#6"),
       `$120$` = c("T", "T", "T", "C", "C", "C"),
       `$100$` = c("T", "C", "C", "T", "T", "C"),
       `$130$` = c("C", "T", "C", "T", "C", "T"),
       `$120$ ` = c("C", "C", "T", "C", "T", "T"),
       `ATE` = c("$-15$", "$+15$", "$+5$", "$-5$", "$-15$", "$+15$")) %>%
       
  # Then, we use the kable function to make it pretty
  
  kable() %>%
  kable_styling(bootstrap_options = "striped",
                full_width = FALSE,
                position = "left") %>%
  column_spec(1,
              bold = TRUE,
              border_right = TRUE)
```

Here, permutation #1 is what we actually observed, and permutations #2-6 are all the possible rearrangement of the two "treatment" labels and the two "control" labels.  What do we see?  4/6 (67%) of the permutations produce calculated ATEs are at least as large as the effect we actually observed ($-15$), and two of them are the opposite sign!  We therefore should not have much confidence from these data alone that the drug really lowers blood pressure, even though the effect size appeared large.  The moral of the story?  Don't conduct an experiment on only four people!

Let's say that we had a larger experiment.  We can no longer calculate the results of the permutation test by hand, since the number of possible permutations will quickly become very large.  However, you can use R to calculate far more permutations than you can do by hand.  Furthermore, if the number of permutations becomes too much even for your computer, you can take a random sample of all the possible permutations instead; if the random sample is large enough, this will give you a result very close to what you would get if you considered all the permutations.

## Confounding and selection bias

With a large enough sample and randomized assignment, we can get a good estimate of the ATE.  Other assignment mechanisms, however, can make it difficult or impossible to engage in causal inference.

Why is that?  Consider the use of the *perfect doctor* as an assignment mechanism. The perfect doctor knows how each subject will respond to the drug or the control and assigns each subject to the treatment that will most benefit her. The perfect doctor knows this information about a sample of patients: 

```{r true_data_perfect_doctor, echo = FALSE}
# First, we create a tibble with the values we want for the table

tibble(`Subject` = c("Joe", "Mary", "Sally", "Bob", "James", "Susie", "MEAN"),
       `$Y_t(u)$` = c("$130$", "$120$", "$100$", "$115$", "$120$", "$135$", "$120$"),
       `$Y_c(u)$` = c("$115$", "$125$", "$150$", "$125$", "$130$", "$105$", "$125$"),
       `$Y_t(u) - Y_c(u)$` = c("$+15$", "$-5$", "$-50$", "$-10$", "$-10$", "$+30$", "$-5$")) %>%
  
  # Then, we use the kable function to make it pretty
  
  kable() %>%
  kable_styling(bootstrap_options = "striped",
                full_width = FALSE,
                position = "left") %>%
  column_spec(1,
              bold = TRUE,
              border_right = TRUE)
```

Based on this knowledge she would make the following treatment assignments: 

```{r perfect_doctor, echo = FALSE}
# First, we create a tibble with the values we want for the table

tibble(`Subject` = c("Joe", "Mary", "Sally", "Bob", "James", "Susie", "MEAN"),
       `$Y_t(u)$` = c("$?$", "$120$", "$100$", "$115$", "$120$", "$?$", "$113.75$"),
       `$Y_c(u)$` = c("$115$", "$?$", "$?$", "$?$", "$?$", "$105$", "$110$"),
       `$Y_t(u) - Y_c(u)$` = c("$?$", "$?$", "$?$", "$?$", "$?$", "$?$", "$+3.75$")) %>%
  
  # Then, we use the kable function to make it pretty
  
  kable() %>%
  kable_styling(bootstrap_options = "striped",
                full_width = FALSE,
                position = "left") %>%
  column_spec(1,
              bold = TRUE,
              border_right = TRUE)
```

The perfect doctor distorts the averages of both $Y_t(u)$ and $Y_c(u)$ by filtering out poor responses to both the treatment and control. The difference in means thus is distorted in a direction that depends on the details. For instance, a subject like Susie who is harmed by taking the drug would be assigned to the control group by the perfect doctor and thus the negative effect of the drug would be masked.

Therefore, the difference in means is no longer a good estimate of the ATE. In fact, in this case it has the wrong sign! This is not merely a consequence of our small sample: even if the perfect doctor saw a million patients, we could not get a good estimate of the ATE.

This is an extreme example of a problem called **selection bias**.  The perfect doctor is not choosing which patients are receiving the drug at random. Rather, the perfect doctor is making treatment decisions based directly on the patient's potential outcomes.  Whenever the assignment mechanism is correlated with the potential outcomes, we say that there is **confounding**.  Confounding is a problem, since it means that our simple estimate of the ATE is biased.

Thus, if you are interested in causal inference, randomized trials are the best approach. In many circumstances, however, randomized trials are not possible due to ethical or practical concerns. In such scenarios there is by necessity a non-random assignment mechanism.

For instance, let's say you were interested in the effect of college attendence on earnings. People are not randomly assigned to attend college. Rather, people may choose to attend college based on their financial situation, parents' education, and so on.  This can introduce confounding if the assignment mechanism affects future earnings.  For example, if people choose to go to college at higher rates when they are on career paths where a college degree is particularly beneficial  -- or, in RCM language, whose potential outcomes are on average higher under treatment -- that would introduce confounding.

Many statistical methods have been developed for causal inference when there is a non-random assignment mechanism, such as propensity score matching. These methods attempt to correct for the assignment mechanism by finding control units similar to treatment units. What you should not do is naively compare the sample means under treatment and control and assume that is a good estimate of the ATE.  Without randomization, this could be very misleading!

## Internal and external validity

If we have randomized assignment and a large sample, we can be confident that we have a good estimate of the average treatment effect *in that sample*.  We say that the experiment has high **internal validity**: the inferences we are making are likely to reflect the truth about that sample.

However, we may be interested in a population beyond our particular sample.  For example, consider the hypertension drug example discussed above.  Let's say that we ran a randomized trial on 10,000 patients, with 5,000 receiving the new drug and 5,000 in the contol group.  We estimate an average treatment effect of $-5$ units of systolic blood pressure.  Should we recommend this new drug?

The answer to that question depends in part on the **external validity** of the study.  Are the 10,000 people in the study similar to the people for whom we will be making recommendations?  Let's say that the 10,000 study participants were chosen because they all lived near the location that the study took place. That's one example of **selection bias**, where the sample is not a random sample of the population in which we are interested.  Why is that a problem?  Those people may differ systematically from other people, including in ways that affect their response to blood pressure medicine. For example, what if the people in this area are younger on average than most people who take blood pressure medicine?

This concern can be expressed in terms of the assignment mechanism.  People who live far away from the location of the study have a 0% chance of receiving the treatment.  Thus, the study *can't* directly speak to whether the treatment is effective for them; the only way we can make such claims is by making additional assumptions, such as that the non-participants will respond the same on average to the drug as the participants.

The circumstances of the experiment may also affect the external validity of the study.  Perhaps the study participants were all eating a diet that was controlled by the experimenters.  Then, while we have variation in one aspect of the treatment (whether the participants received the blood pressure medicine), we don't have variation in another (diet).  It may be that the blood pressure medicine works when on the controlled diet but doesn't work otherwise.

When dealing with human subjects, there is a particular concern regarding external validity: the [**Hawthorne effect**](https://en.wikipedia.org/wiki/Hawthorne_effect).  When human subjects know that they are part of an experiment, they may change their behavior.

This can lead to a version of the above concern.  Suppose again that the blood pressure drug works when combined with a controlled diet, say a low-salt diet, but it doesn't work for those who have a high-salt diet.  If the subjects in the experiment are paying greater attention to their blood pressure than they otherwise would and thus are eating lower sodium diets than other people, one could make the mistaken inference that it was the blood pressure drug alone that led to a decrease in blood pressure, and not the combination of the drug and the experimental subjects' changing behavior.

## Survey research and external validity

<!-- Should this be merged with the sampling chapter? -->

In political science and other social sciences, survey research poses some particular problems of external validity.  Let's say that you want to measure presidential approval among all U.S. voters.  How can you make sure that the sample you collect is representative of your sampling frame?  (In the language of survey sampling, the "sampling frame" is the population you want to study: here, voters in the United States.)

The most natural approach is to engage in a simple random sample: each U.S. voter has the same probability of being in the survey.  This would ensure that one's sample was on average representative of the population, although there will still be some sampling error because the sample is finite.  About 139 million people voted in the 2016 general election,^[See [here](http://www.electproject.org/2016g).] so if one wanted a sample of 1,390 voters, one would like a sampling method that gives each voter an equivalent 0.001% chance of being surveyed.

But of course, there's no method of assuring that each voter has an equivalent chance of being in your survey!  Let's say that you decided to conduct your survey via random digit dialing, where you call possible phone numbers at random and give your survey to those who answer.  There are many possible sources of bias in this measure:

- Some voters may not have telephone numbers
- Some voters may be more likely to answer the phone and take your survey than others
- Some people may lie or misremember about whether they voted in the 2016 general election
- More subtly, respondents' presidential approval may be affected by other questions you ask in the survey. (Perhaps this is what you are studying, for instance if you were testing the effect of a particular message on presidential approval, but if not, you need to be careful in interpreting your results.)

And these are just the most obvious!  If you take a course in survey research, you'll learn more about the ways in which pollsters attempt to deal with these problems.

With the growing popularity of survey experiments in social research, it is always worth considering how the sampling was conducted.

## Conclusion

The causal effect of a treatment on a single unit at a point in time is the difference between the outcome variable with the treatment and without the treatment. The Fundamental Problem of Causal Inference is that it is impossible to observe the causal effect on a single unit. You either take the aspirin now or you don't. As a consequence, assumptions must be made in order to estimate the missing counterfactuals.

## References

<!-- Need to integrate this with the rest of the book -->

Cox, D. R. (1958). *Planning of Experiments.* Wiley.

Holland, Paul W. (1986). "Statistics and Causal Inference." *J. Amer. Statist. Assoc.* 81 (396): 945–960. doi:10.1080/01621459.1986.10478354.

Neyman, Jerzy. (1923). *Sur les applications de la theorie des probabilites aux experiences agricoles: Essai des principes.* Master's Thesis. Excerpts reprinted in English, Statistical Science, Vol. 5, pp. 463–472. (D. M. Dabrowska, and T. P. Speed, Translators.)

Rubin, Donald (1974). "Estimating Causal Effects of Treatments in Randomized and Nonrandomized Studies." *J. Educ. Psychol.* 66 (5): 688–701 [p. 689]. doi:10.1037/h0037350.

Rubin, Donald (2005). "Causal Inference Using Potential Outcomes." *J. Amer. Statist. Assoc.* 100 (469): 322–331. doi:10.1198/016214504000001880.
