# (APPENDIX) Appendix {-}

# Rubin Causal Model {#rubin-causal-model}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(gt)
library(tidyverse)
```

The Rubin Causal Model (RCM) is an approach to the statistical analysis of cause and effect based on the framework of potential outcomes, named after Donald Rubin.^[This chapter remixes some material from [Rubin Causal Model (Wikipedia)](https://en.wikipedia.org/wiki/Rubin_causal_model). <br /> The name "Rubin Causal Model" was first coined by Paul W. Holland (1986).]  

<!-- AR: Perhaps the credit to Wikipedia should be added to the attributions page
rather than here.-->

## What is a causal effect?

The Rubin Causal Model (RCM) is based on the idea of **potential outcomes.**  For example, let's say you took blood pressure medicine and then measure your blood pressure. To measure the causal effect of taking the medicine, we need to compare the outcome for you in one alternative future (where you took the medicine) to another (where you did not). Since it is impossible to see both potential outcomes at once, one of the potential outcomes is always missing. This dilemma is the "fundamental problem of causal inference."

In most circumstances, we are interested in comparing two futures, one generally termed "treatment" and the other "control." These labels are somewhat arbitrary. The difference between the potential outcome under treatment and the potential outcome under control is called a "causal effect" or a "treatment effect."  The scenario that didn't actually happen, and thus that we didn't observe, is called a "counterfactual."

Thus, according to the RCM, the **causal effect** of your taking or not taking the medicine is the *difference* between what your blood pressure would have been under "treatment" (taking the medicine) and "control" (not taking the medicine). If your systolic blood pressure would have been 140 without the medicine and was 120 with it, then the causal effect of taking the medicine is a 20-point reduction in your systolic blood pressure.

### Everything is a missing data problem

<!-- AR: This perhaps could be introduced even earlier, before the students are
assigned the RCM chapter.  Could be a motivating idea for the whole book. -->

Before we discuss the RCM in more detail, it's worth discussing in general the problem of **missing data**.  Why is that?  When you don't observe certain counterfactuals, that's just one example of a missing piece of data.  Missing data problems arise whenever you want to make statistical inferences.

For example, let's say that before you test a new blood pressure medicine, you want to know the average systolic blood pressure of United States adults.  At first, this seems like a much easier problem---there's nothing causal here!  If you knew the true values, you could build a dataset like this: 


```{r, echo = FALSE}
# First, we create a tibble with the values we want for the table

tibble(subject = c("Person 1", "Person 2", "Person 3", "...", "Person N"),
       `Blood Pressure` = c("130", "110", "115", "...", "140")) %>%
  
  # Then, we use the gt function to make it pretty
  
  gt() %>%
  cols_label(subject = md("**Person**")) %>%
  tab_style(cell_borders(sides = "right"),
            location = cells_body(columns = vars(subject))) %>%
  tab_style(cell_text(weight = "bold"),
            location = cells_body(columns = vars(subject)))
```

Then, your answer is simply the average of all the values. 

But do we have this table?  No!  What we actually have is this:

```{r, echo = FALSE}
# First, we create a tibble with the values we want for the table

tibble(subject = c("Person 1", "Person 2", "Person 3", "...", "Person N"),
       `Blood Pressure` = c("?", "?", "?", "...", "?")) %>%
  
  # Then, we use the gt function to make it pretty
  
  gt() %>%
  cols_label(subject = md("**Person**")) %>%
  tab_style(cell_borders(sides = "right"),
            location = cells_body(columns = vars(subject))) %>%
  tab_style(cell_text(weight = "bold"),
            location = cells_body(columns = vars(subject)))
```

That is, we have a lot of *missing data*.

But maybe we could survey 1,000 people, and ask them to report their blood pressures:

```{r, echo = FALSE}
# First, we create a tibble with the values we want for the table

tibble(subject = c("Respondent 1", "Respondent 2", "Respondent 3", "...", "Respondent 1,000"),
       `Blood Pressure` = c("130", "110", "115", "...", "140")) %>%
  
  # Then, we use the gt function to make it pretty
  
  gt() %>%
  cols_label(subject = md("**Respondent**")) %>%
  tab_style(cell_borders(sides = "right"),
            location = cells_body(columns = vars(subject))) %>%
  tab_style(cell_text(weight = "bold"),
            location = cells_body(columns = vars(subject)))
```

At least then we could have an average blood pressure for the respondents in our sample, right?

Not so fast! It is probably not fair to assume that every blood pressure measurement is perfectly accurate.  Some survey respondents may misremember or round the values, or use measurements from a doctor's visit long ago.  (How many American adults own blood pressure monitors?)  Furthermore, blood pressure monitors aren't 100% accurate.

One's blood pressure reading also isn't a single, constant number, even if it could be perfectly measured: it depends on the time of day, whether one exercised or ate before taking the measurement, and so on.

So what we really should do is think of each measurement as being drawn from a distribution of potential values for each person; we could construct a *confidence interval* around each of the values to try to reflect these sources of error.  Let's say that a fair CI is 15 points in either direction.  (It may actually be higher; this is just for illustration.)  Then our table really looks something like this:

```{r, echo = FALSE}
# First, we create a tibble with the values we want for the table

tibble(subject = c("Respondent 1", "Respondent 2", "Respondent 3", "...", "Respondent 1,000"),
       `Blood Pressure` = c("130 (115, 145)", "110 (95, 125)", "115 (100, 130)", "...", "140 (125, 155)")) %>%
  
  # Then, we use the gt function to make it pretty
  
  gt() %>%
  cols_label(subject = md("**Respondent**")) %>%
  tab_style(cell_borders(sides = "right"),
            location = cells_body(columns = vars(subject))) %>%
  tab_style(cell_text(weight = "bold"),
            location = cells_body(columns = vars(subject)))
```

OK, so with all this information, we can compute an average blood pressure reading for our respondents, and we can also get a sense of how confident we should be in that average.  With a big enough sample, we can probably get a pretty good measure of the average blood pressure for that sample.  Have we solved our problem?  That is, can we now assume that our estimate of the average blood pressure in our sample is a good estimate of the average blood pressure in the U.S. adult population?

<!-- AR: Talking about sampling here so we can reference it later -->

Well, let's see how many of the missing values we have filled in to our table of the population:

```{r, echo = FALSE}
# First, we create a tibble with the values we want for the table

tibble(subject = c("Person 1", "Person 2", "Person 3", "...", "Person N"),
       `Blood Pressure` = c("?", "?", "?", "...", "?")) %>%
  
  # Then, we use the gt function to make it pretty
  
  gt() %>%
  cols_label(subject = md("**Person**")) %>%
  tab_style(cell_borders(sides = "right"),
            location = cells_body(columns = vars(subject))) %>%
  tab_style(cell_text(weight = "bold"),
            location = cells_body(columns = vars(subject)))
```

Alas, since the U.S. population is over 300 million, we can't even see our 1,000 respondents!  If we want to assume that our estimate for our respondents is a good estimate for the population, we need to assume that their blood pressures are, on average, the same as the blood pressures of the people we *didn't* survey.

For that to be true, it means that the *sampling mechanism* can't be related to someone's blood pressure.  For example, let's say that we surveyed 1,000 Harvard undergraduates.  Do we think that Harvard undergraduates have average blood pressure? No! Harvard undergraduates are younger than average, and younger people have lower blood pressures.

So we should *randomly* sample the population.  Then, there won't be any systematic relationship between being chosen for our sample and blood pressure.  (There may be a relationship by chance, but taking a bigger sample reduces that possibility.)

But even randomly choosing whom to survey may not totally solve our problem.  Some people may not fill out our survey, after all.  So our table for our survey may look like this:

```{r, echo = FALSE}
# First, we create a tibble with the values we want for the table

tibble(subject = c("Respondent 1", "Respondent 2", "Respondent 3", "...", "Respondent 1,000"),
       `Blood Pressure` = c("130", "?", "115", "...", "140")) %>%
  
  # Then, we use the gt function to make it pretty
  
  gt() %>%
  cols_label(subject = md("**Respondent**")) %>%
  tab_style(cell_borders(sides = "right"),
            location = cells_body(columns = vars(subject))) %>%
  tab_style(cell_text(weight = "bold"),
            location = cells_body(columns = vars(subject)))
```

So not only do we have all the ? values in the population table, we also have a ? for one of our survey respondents (Respondent 2).  Can we just fill in the ? for Respondent 2 with the average of the other respondents?  Only if the factors that led Respondent 2 not to answer the survey had nothing to do with blood pressure.  Let's say that poorer respondents were less likely to fill out the survey, for example.  If poorer respondents also had higher blood pressure on average, our estimate of the average blood pressure *among those who answered the survey* will be biased downwards as an estimate for the population, even if we surveyed people at random. 

So missing data is not just a problem for estimating causal effects.  The mechanisms by which you find out some things and don't find out other things are critical to understand when trying to make statistical inferences.

### Potential outcomes: introduction to the Rubin Table

Suppose that Joe is participating in an FDA test for a new hypertension drug. If we were omniscient, we would know the outcomes for Joe under both treatment (the new drug) and control (either no treatment or the current standard treatment). The causal effect, or treatment effect, is the difference between these two potential outcomes. We can present this with the following table of potential outcomes, which we will call a Rubin Table:

<!-- AR: Should we stick with this notation? -->

```{r, echo = FALSE}
# First, we create a tibble with the values we want for the table

tibble(subject = "Joe",
       ytreat = "130",
       ycontrol = "135",
       ydiff = "-5") %>%
  
  # Then, we use the gt function to make it pretty
  
  gt() %>%
  cols_label(subject = md("$$\\mathbf{Subject}$$"),
                ytreat = md("$$Y_t(u)$$"),
                ycontrol = md("$$Y_c(u)$$"),
                ydiff = md("$$Y_t(u) - Y_c(u)$$")) %>%
  tab_style(cell_borders(sides = "right"),
            location = cells_body(columns = vars(subject))) %>%
  tab_style(cell_text(weight = "bold"),
            location = cells_body(columns = vars(subject)))
```


$Y_{t}(u)$ is Joe's systolic blood pressure if he takes the new pill. In general, this notation expresses the potential outcome which results from a treatment, $t$, on a unit, $u$. Similarly, $Y_{c}(u)$ is the effect of a different treatment, $c$ or control, on a unit, $u$. In this case, $Y_{c}(u)$ is Joe's blood pressure if he doesn't take the pill. $Y_{t}(u)-Y_{c}(u)$ is the causal effect of taking the new drug.  Thus, Joe would have a blood pressure of $130$ if he took the pill and $135$ if he did not; the causal effect of taking the pill is $-5$.  Joe should take the pill!

From this table we only know the causal effect on Joe. Everyone else in the study might have an increase in blood pressure if they take the pill. However, regardless of what the causal effect is for the other subjects, the causal effect for Joe is lower blood pressure, relative to what his blood pressure would have been if he had not taken the pill.

### No causation without manipulation

In order for a potential outcome to make sense, it must be possible, at least *a priori*. For example, if there is no way for Joe, under any circumstance, to obtain the new drug, then $Y_{t}(u)$ is impossible for him. It can never happen. And if $Y_{t}(u)$ can never be observed, even in theory, then the causal effect of treatment on Joe's blood pressure is not defined. 

The causal effect of new drug is well defined because it is the simple difference of two potential outcomes, both of which might happen. In this case, we (or something else) can manipulate the world, at least conceptually, so that it is possible that one thing or a different thing might happen.

This definition of causal effects becomes much more problematic if there is no way for one of the potential outcomes to happen, ever. For example, what is the causal effect of Joe's height on his weight? Naively, this seems similar to our other examples. We just need to compare two potential outcomes: what would Joe's weight be under the treatment (where treatment is defined as being 3 inches taller) and what would Joe's weight be under the control (where control is defined as his current height).

A moment's reflection highlights the problem: we can't increase Joe's height. There is no way to observe, even conceptually, what Joe's weight would be if he were taller because there is no way to make him taller. We can't manipulate Joe's height, so it makes no sense to investigate the causal effect of height on weight. Hence the slogan: **No causation without manipulation.**

### Estimands

Let's look again at our basic Rubin Table for Joe:

```{r, echo = FALSE}
# First, we create a tibble with the values we want for the table

tibble(subject = "Joe",
       ytreat = "130",
       ycontrol = "135",
       ydiff = "-5") %>%
  
  # Then, we use the gt function to make it pretty
  
  gt() %>%
  cols_label(subject = md("$$\\mathbf{Subject}$$"),
                ytreat = md("$$Y_t(u)$$"),
                ycontrol = md("$$Y_c(u)$$"),
                ydiff = md("$$Y_t(u) - Y_c(u)$$")) %>%
  tab_style(cell_borders(sides = "right"),
            location = cells_body(columns = vars(subject))) %>%
  tab_style(cell_text(weight = "bold"),
            location = cells_body(columns = vars(subject)))
```

We have Joe's potential outcome under treatment, Joe's potential outcome under control, and the difference between the two.  The difference is often called the "treatment effect."  But it is in fact only one possible **estimand**.  An estimand is some quantity in the real world that we are trying to measure.  Note that we could also estimate the ratio of potential outcomes for Joe:

```{r, echo = FALSE}
# First, we create a tibble with the values we want for the table

tibble(subject = "Joe",
       ytreat = "130",
       ycontrol = "135",
       yratio = "26/27") %>%
  
  # Then, we use the gt function to make it pretty
  
  gt() %>%
  cols_label(subject = md("$$\\mathbf{Subject}$$"),
                ytreat = md("$$Y_t(u)$$"),
                ycontrol = md("$$Y_c(u)$$"),
                yratio = md("$$Y_t(u) / Y_c(u)$$")) %>%
  tab_style(cell_borders(sides = "right"),
            location = cells_body(columns = vars(subject))) %>%
  tab_style(cell_text(weight = "bold"),
            location = cells_body(columns = vars(subject)))
```

Or the percentage change in potential outcomes:

```{r, echo = FALSE}
# First, we create a tibble with the values we want for the table

tibble(subject = "Joe",
       ytreat = "130",
       ycontrol = "135",
       yperc = "-3.7%") %>%
  
  # Then, we use the gt function to make it pretty
  
  gt() %>%
  cols_label(subject = md("$$\\mathbf{Subject}$$"),
                ytreat = md("$$Y_t(u)$$"),
                ycontrol = md("$$Y_c(u)$$"),
                yperc = md("$$(Y_t(u) - Y_c(u)) * 100 / Y_c(u)$$")) %>%
  tab_style(cell_borders(sides = "right"),
            location = cells_body(columns = vars(subject))) %>%
  tab_style(cell_text(weight = "bold"),
            location = cells_body(columns = vars(subject)))
```

Now, let's consider a larger sample of patients: 

```{r, echo = FALSE}
# First, we create a tibble with the values we want for the table

tibble(subject = c("Joe", "Mary", "Sally", "Bob", "James", "MEAN"),
       ytreat = c("130", "130", "130", "140", "145", "135"),
       ycontrol = c("135", "145", "145", "150", "140", "143"),
       ydiff = c("-5", "-15", "-15", "-10", "+5", "-8")) %>%
  
  # Then, we use the gt function to make it pretty
  
  gt() %>%
  cols_label(subject = md("$$\\mathbf{Subject}$$"),
                ytreat = md("$$Y_t(u)$$"),
                ycontrol = md("$$Y_c(u)$$"),
                ydiff = md("$$Y_t(u) - Y_c(u)$$")) %>%
  tab_style(cell_borders(sides = "right"),
            location = cells_body(columns = vars(subject))) %>%
  tab_style(cell_text(weight = "bold"),
            location = cells_body(columns = vars(subject)))
```

From this Rubin Table, there are many possible estimands.  Here are just a sampling:

- One potential outcome for one person, such as Joe's potential outcome under treatment ($130$).  
- A causal effect for one person, such as for Mary.  This is the difference between the potential outcomes, which we have provided as its own column ($130 - 145 = -15$).
- The most positive causal effect.  Here, that is $+5$, from Bob ($145 - 140 = +5$).
- The most negative causal effect.  Here, that is $-15$, from either Mary or Sally (both $130 -145 = -15$).
- The median causal effect ($-10$).
- The median percentage change.  To do this, calculate the percentage change for each person.  You'll get 5 percentages: $-3.7\%$, $-10.3\%$, $-10.3\%$, $-6.7\%$, and $3.6\%$.  The median is $-6.7\%$.
- The total number of people for whom the causal effect is positive: $4$.
- And so on. There are a lot of things one might care about!

<!-- AR: the ATE is just another estimand--and a particularly important one--so
I think it should be mentioned here.  We can talk about calculating it from
actual data later.-->

<!-- AR: Calling this the "average treatment effect" rather than "average causal
effect" throughout -->

One very common estimand has its own name, the **average treatment effect**.  The average treatment effect (often abbreviated **ATE**) is the mean of all the causal effects.  Here, the mean is $-8$.

All of these can be derived if you know the full Rubin Table, but in real life only one of these potential outcomes can actually be observed.  Still, the **estimand** is the value that you would calculate, if you had the full Rubin Table.

Of course, we don't have the full Rubin Table. It is impossible, by definition, to observe the effects of more than one treatment on a single subject. Joe cannot both take the pill and not take the pill at the same time! Therefore, the data we actually observe would look something like this: 

```{r, echo = FALSE}
# First, we create a tibble with the values we want for the table

tibble(subject = c("Joe", "Mary", "Sally", "Bob", "James", "MEAN"),
       ytreat = c("130", "120", "?", "?", "115", "121.67"),
       ycontrol = c("?", "?", "125", "130", "?", "127.5"),
       ydiff = c("?", "?", "?", "?", "?", "-5.83")) %>%
  
  # Then, we use the gt function to make it pretty
  
  gt() %>%
  cols_label(subject = md("$$\\mathbf{Subject}$$"),
                ytreat = md("$$Y_t(u)$$"),
                ycontrol = md("$$Y_c(u)$$"),
                ydiff = md("$$Y_t(u) - Y_c(u)$$")) %>%
  tab_style(cell_borders(sides = "right"),
            location = cells_body(columns = vars(subject))) %>%
  tab_style(cell_text(weight = "bold"),
            location = cells_body(columns = vars(subject)))
```

Question marks are responses that could not be observed. The **Fundamental Problem of Causal Inference**^[Holland 1986.] is that directly observing unit-level causal effects is impossible.

### The infinite Rubin Table: many kinds of missing data

While this is what our *observed* Rubin Table may look like, just filling in these question marks likely wouldn't be enough to answer our scientific question of interest.  We are likely interested in whether the new blood pressure drug works *in general*, not just for these six people. Just as when we were considering estimating the average blood pressure of every adult in the U.S., we'll need to think about whether our *sample* is representative of the population.  Thus, we usually have (at least) two main sources of missing data:

1. For the units in our sample, we only see one potential outcome
1. For the units outside our sample, we see *no* potential outcomes

Furthermore, even if we did want to know just about these six people, do we only care about their outcomes this year?  No!  We also care about Joe's potential outcomes one year from now, two years from now, and so on.

So our full Rubin Table includes people we know (Joe) and people we don't (for example, Tyrone), both now and in the future:

```{r, echo = FALSE}
# First, we create a tibble with the values we want for the table

tibble(subject = c("Joe this year", "Joe next year", "Joe two years from now", "Tyrone this year", "Tyrone next year", "Tyrone two years from now", "...", "Person n at time t"),
       ytreat = c("130", "?", "?", "?", "?", "?", "?", "?"),
       ycontrol = c("?", "?", "?", "?", "?", "?", "?", "?"),
       ydiff = c("?", "?", "?", "?", "?", "?", "?", "?")) %>%
  
  # Then, we use the gt function to make it pretty
  
  gt() %>%
  cols_label(subject = md("$$\\mathbf{Subject}$$"),
                ytreat = md("$$Y_t(u)$$"),
                ycontrol = md("$$Y_c(u)$$"),
                ydiff = md("$$Y_t(u) - Y_c(u)$$")) %>%
  tab_style(cell_borders(sides = "right"),
            location = cells_body(columns = vars(subject))) %>%
  tab_style(cell_text(weight = "bold"),
            location = cells_body(columns = vars(subject)))
```

In fact, because time is continuous, there is a row for Joe now, Joe one second from now, Joe one day from now and so on. The Rubin Table extends downward forever. Thus, in order to estimate any causal effect, we need assumptions, so we aren't dealing with an infinite table.

The most obvious way to eliminate some rows from the table is to assume the causal effect for Joe now is the same as all the ones for Joe in the future. Is that plausible? Sort of. Joe now and Joe in one second are pretty similar! Joe now and Joe in 30 years are less so.  Unfortunately, there's no magic way to get a good estimate of every missing value in the infinite Rubin Table!

Not only can we extend the Rubin Table by adding people not in our sample to the rows, but we can also add additional treatments to the columns.  Let's go back to the original six people in our sample. What if instead of testing one blood pressure drug, we wanted to test two?  We'll call the original treatment $t$ and the new treatment $t'$.

```{r, echo = FALSE}
# First, we create a tibble with the values we want for the table

tibble(subject = c("Joe", "Mary", "Sally", "Bob", "James", "MEAN"),
       ytreat = c("130", "120", "?", "?", "115", "121.67"),
       ytreat2 = c("?", "?", "?", "?", "?", "?"),
       ycontrol = c("?", "?", "125", "130", "?", "127.5")) %>%
  
  # Then, we use the gt function to make it pretty
  
  gt() %>%
  cols_label(subject = md("$$\\mathbf{Subject}$$"),
             ytreat = md("$$Y_t(u)$$"),
             ytreat2 = md("$$Y_{t'}(u)$$"),
             ycontrol = md("$$Y_c(u)$$")) %>%
  tab_style(cell_borders(sides = "right"),
            location = cells_body(columns = vars(subject))) %>%
  tab_style(cell_text(weight = "bold"),
            location = cells_body(columns = vars(subject)))
```

Note that for Joe, we now have three causal effects we can estimate: the difference between the original treatment and the new treatment, the difference between the original treatment and control, and the difference between the new treatment and control.  And that's just when looking at differences!  Recall there are many other potential estimands we could be interested in, such as ratios, percent changes, and so on.

Even if you have just one medication that you are testing, there still could be multiple treatments.  For example, the patients could take a different number of pills:

```{r, echo = FALSE}
# First, we create a tibble with the values we want for the table

tibble(subject = c("Joe", "Mary", "Sally", "Bob", "James", "MEAN"),
       ytreat = c("130", "120", "?", "?", "115", "121.67"),
       ytreat2 = c("?", "?", "?", "?", "?", "?"),
       ytreat3 = c("?", "?", "?", "?", "?", "?"),
       ytreat4 = c("?", "?", "?", "?", "?", "?"),
       ytreat5 = c("?", "?", "?", "?", "?", "?"),
       ycontrol = c("?", "?", "125", "130", "?", "127.5")) %>%
  
  # Then, we use the gt function to make it pretty
  
  gt() %>%
  cols_label(subject = md("$$\\mathbf{Subject}$$"),
             ytreat = md("$$Y_{\\text{1 pill}}(u)$$"),
             ytreat2 = md("$$Y_{\\text{2 pills}}(u)$$"),
             ytreat3 = md("$$Y_{\\text{3 pills}}(u)$$"),
             ytreat4 = md("$$Y_{\\text{4 pills}}(u)$$"),
             ytreat5 = md("$$Y_{\\text{5 pills}}(u)$$"),
             ycontrol = md("$$Y_c(u)$$")) %>%
  tab_style(cell_borders(sides = "right"),
            location = cells_body(columns = vars(subject))) %>%
  tab_style(cell_text(weight = "bold"),
            location = cells_body(columns = vars(subject)))
```

Again, there are many possible estimands.  It's worth noting when the treatments have some numeric relationship to each other, though, there may be reasonable assumptions we can use to help fill in the missing values---but we aren't there yet!

Instead of considering the treatment in terms of discrete pills, we could also consider different doses: 1000 mg, 1001 mg, and so on:

```{r, echo = FALSE}
# First, we create a tibble with the values we want for the table

tibble(subject = c("Joe", "Mary", "Sally", "Bob", "James", "MEAN"),
       ytreat = c("130", "120", "?", "?", "115", "121.67"),
       ytreat1001 = c("?", "?", "?", "?", "?", "?"),
       ytreat1002 = c("?", "?", "?", "?", "?", "?"),
       yellipsis = c("?", "?", "?", "?", "?", "?"),
       ycontrol = c("?", "?", "125", "130", "?", "127.5")) %>%
  
  # Then, we use the gt function to make it pretty
  
  gt() %>%
  cols_label(subject = md("$$\\mathbf{Subject}$$"),
             ytreat = md("$$Y_{\\text{1000 mg}}(u)$$"),
             ytreat1001 = md("$$Y_{\\text{1001 mg}}(u)$$"),
             ytreat1002 = md("$$Y_{\\text{1002 mg}}(u)$$"),
             yellipsis = md("$$Y_{\text{...}}(u)$$"),
             ycontrol = md("$$Y_c(u)$$")) %>%
  tab_style(cell_borders(sides = "right"),
            location = cells_body(columns = vars(subject))) %>%
  tab_style(cell_text(weight = "bold"),
            location = cells_body(columns = vars(subject)))
```

Indeed, there are an infinite number of possible treatments. The Rubin Table extends to the right forever. Again, assumptions come to our rescue. Or rather, we just throw up our hands and only try to estimate a few things.  This is why it is crucial to define one's estimand precisely: if we are interested in the difference in potential outcomes between one 1000 mg pill of the new drug versus control, we can ignore all the other possible columns in the infinite Rubin Table.

Thus, whenever you are considering a causal question, the best way to think about it is to start with the infinite Rubin Table.  First we throw out the rows we think are duplicates (such as all the observations for Joe one second from now, two seconds from now, etc.) or that are outside the scope of what we are interested in for now (maybe we don't care about outcomes 30 years in the future for this study).  Second, we throw out the columns that we don't care about, which are all the possible treatments we aren't considering.  Finally, we define precisely---in terms of potential outcomes---our estimand.  It may be something simple, such as the average treatment effect, or something more complex.  Once we have done these steps, we can start thinking about how to fill in the question marks.  But remember that the infinite Rubin Table is always there, and you should be conscious of which rows and columns you are throwing out!


## How do we fill in the missing values?

<!-- All the above starts with the introduction of the 6 person example. Are we ready, at this point, to talk about estimation? No! Right now, I think it goes a little quickly. Too early to start talking about ATE. Instead: -->

<!-- Talk about how you fill in the question marks, but going back to the 6 people and just the treatment/control. Start case is that you can't fill them in. You can never know! Next step is that, even if you could know, everyone is different, so every causal effect is different. But not much you can do with that easier. So, we start making assumptions. And "assumption" means a "model," and all models have parameters.  -->

<!-- One model might be that the causal effect is the same for everyone. There is a single parameter, tau, which we then estimate. Once we have an estimate, we can fill in the Rubin Table because, knowing it, we know what the unseen potential outcome is for each person. This provides another opportunity to introduce the concept of estimand, which is tau in this case. -->
  
<!-- A second model might be that the causal effect is the same within categories. That is, it is tau_1 for men and tau_2 for women.  A third model might be . . . The key concept is that we can't make any progress unless we make some assumptions. That is an inescapable result of the fundamental problem of causal inference. So we make some assumptions, which give us some models. -->



<!-- DK: And now, finally, we can estimate ATE, as we do below. But we should highlight that this is just one model. A different model, like tau varies by gender, would require a different thing to estimate. ATE is not really special. It is just one model which allows us to fill in the question marks. So, we do this twice. Filling in the question marks differently each time. And this highlights the dilemma. We can never really know what Joe's outcome under control would have been. In fact, with just two different models, we have two different guesses. -->

<!-- And now revisit the concept of estimand. Because things get a bit circular here! Let's say that the estimand we really care about is --- What is the lowest blood pressure after treatment? That is not an unreasonable thing to want to know. But we are not, under either of our two models, estimating that directly! Instead, we are estimating tau (or tau_1 and tau_2) and then using those estimates to populate the table. Then, with the complete table, we can get the number we really care about. We use one estimand (tau) to get another estimand (lowest blood pressure under treatment), rather than going for the second estimand immediately.  This is important. -->

<!-- DK: Leftover discussion. Not sure it still matters.  How do we fill in the question marks? One answer: Take the mean of the column. Consider the control group. You have their blood pressure. If I find a new person, whose blood pressure you don't know, what would you guess? The mean, obviously. Fill those in. Now you can estimate the causal effect for each person, and the average causal effect. Cool! Is that a good model? No! It sucks, because you have information for each person you are not using: their blood pressure under the other outcome. How to use it? Make a new model in which there is a single parameter, tau, which is the constant treatment effect for everyone. Estimate that. Fill in the ?.  -->

<!-- Only at the very end of this discussion to we mention that, even though there are a lot of estimands that one might be interested in, some, like ATE, are very popular. Then, go through ATE as an example. -->

<!-- DK: At this point, we have Need to add uncertainty into the mix, both the uncertainty in estimating t and the uncertainty in predicting the ? for specific individuals. Maybe mention measurement error? Evn though we pretend that the potential outcomes are fixed (a long with pre-treatment covariates), we never know them exactly. Someone's blood pressure is never exactly 130. There is uncertainty, perhaps best considered as a PDF. And different people might have different amounts of uncertainty.  -->

<!-- With that clear --- and is there a way to add a little normal distribution density to a RT? or something like that? --- we can go on to talk about the uncertainty in the unobserved potential outcome. Indeed, even if we could magically see it, there would still be a PDF because of measurement error. That is the minimum amount of uncertainty we deal with. There are two sources: model uncertainty, which goes away with big N, and individual variation which does not. Fill in the ? with those confidence intervals, the second much wider than the first. In explaining this, you may assume that someone has read the probability and Bayesian chapter of the book. -->

Let's go back to a simple version of the Rubin Table: six people and one binary treatment variable. One way we could solve this problem is by filling in the missing values for treatment and control with the mean values.  That is, let's say we wanted to know what Joe's outcome would have been under control.  We then just plug in the observed mean ($127.5$) of the control observations for Joe's outcome:

```{r, echo = FALSE}
# First, we create a tibble with the values we want for the table

tibble(subject = c("Joe", "Mary", "Sally", "Bob", "James", "MEAN"),
       ytreat = c("130", "120", "121.67", "121.67", "115", "121.67"),
       ycontrol = c("127.5", "127.5", "125", "130", "127.5", "127.5"),
       ydiff = c("2.5", "-7.5", "-3.33", "-8.33", "-12.5", "-5.83")) %>%
  
  # Then, we use the gt function to make it pretty
  
  gt() %>%
  cols_label(subject = md("$$\\mathbf{Subject}$$"),
                ytreat = md("$$Y_t(u)$$"),
                ycontrol = md("$$Y_c(u)$$"),
                ydiff = md("$$Y_t(u) - Y_c(u)$$")) %>%
  tab_style(cell_borders(sides = "right"),
            location = cells_body(columns = vars(subject))) %>%
  tab_style(cell_text(weight = "bold"),
            location = cells_body(columns = vars(subject)))
```

We see that this method doesn't change our estimate of the treatment mean ($121.67$), the control mean ($127.5$), or the mean difference between treatment and control ($-5.83$).  However, it does allow us to make predictions for individual observations.  For example, now we predict that Joe's systolic blood pressure under control would have been $127.5$.

But how confident should we be that Joe's outcome under control is $127.5$?  Not very!  Even if plugging in the average outcome value for the control observations is a good prediction for Joe (and it may well not be), we should recognize that we have a great deal of *uncertainty* about that value.  In Chapter \@ref(confidence-intervals), we show how to construct a *confidence interval* around a quantity of interest; later in this chapter we'll discuss *permutation tests*, a way of quantifying uncertainty for randomized experiences.

However, with only two observations in the control group, we can simply note that we've observed values under control as low as $125$ and as high as $130$, so we don't have good reason to think that Joe's value will be precisely $127.5$.  Here's the same table, but with some makeshift confidence intervals, taken from the lowest and highest values observed under treatment and control:

```{r, echo = FALSE}
# First, we create a tibble with the values we want for the table

tibble(subject = c("Joe", "Mary", "Sally", "Bob", "James", "MEAN"),
       ytreat = c("130", "120", "121.67 (115, 130)", "121.67 (115, 130)", "115", "121.67"),
       ycontrol = c("127.5 (125, 130)", "127.5 (125, 130)", "125", "130", "127.5 (125, 130)", "127.5"),
       ydiff = c("2.5", "-7.5", "-3.33", "-8.33", "-12.5", "-5.83")) %>%
  
  # Then, we use the gt function to make it pretty
  
  gt() %>%
  cols_label(subject = md("$$\\mathbf{Subject}$$"),
                ytreat = md("$$Y_t(u)$$"),
                ycontrol = md("$$Y_c(u)$$"),
                ydiff = md("$$Y_t(u) - Y_c(u)$$")) %>%
  tab_style(cell_borders(sides = "right"),
            location = cells_body(columns = vars(subject))) %>%
  tab_style(cell_text(weight = "bold"),
            location = cells_body(columns = vars(subject)))
```


We may wonder still whether this is a good estimate for Joe.  After all, we are guessing here that the treatment is likely to *raise* his blood pressure, given that his observed blood pressure under treatment was still higher than the average blood pressure under control ($130 > 127.5$).

A way to make a better prediction for Joe is to use an estimate of the average treatment effect.  If we knew that the drug lowered blood pressure on average, that would help us make decisions even if we can't observe the causal effect of the drug on individual people (because we don't observe counterfactuals).

The most natural place to start is with the difference in the sample mean outcomes between treatment and control, $-5.83$.  Note that unlike earlier, this is an *estimate* of the ATE, not the ATE itself.  Why?  Because we have missing data!  For each unit, we only observe one of the potential outcomes.

Thus, another way we can fill in our missing values is by adding our estimate of the ATE to the observed value under control (for units in the control group) or by subtracting our estimate of the ATE from the observed value under treatment (for units in the treatment group), like so:

```{r, echo = FALSE}
# First, we create a tibble with the values we want for the table

tibble(subject = c("Joe", "Mary", "Sally", "Bob", "James", "MEAN"),
       ytreat = c("130", "120", "119.17", "124.17", "115", "121.67"),
       ycontrol = c("135.83", "125.83", "125", "130", "120.83", "127.5"),
       ydiff = c("-5.83", "-5.83", "-5.83", "-5.83", "-5.83", "-5.83")) %>%
  
  # Then, we use the gt function to make it pretty
  
  gt() %>%
  cols_label(subject = md("$$\\mathbf{Subject}$$"),
                ytreat = md("$$Y_t(u)$$"),
                ycontrol = md("$$Y_c(u)$$"),
                ydiff = md("$$Y_t(u) - Y_c(u)$$")) %>%
  tab_style(cell_borders(sides = "right"),
            location = cells_body(columns = vars(subject))) %>%
  tab_style(cell_text(weight = "bold"),
            location = cells_body(columns = vars(subject)))
```

Do we think that if Joe didn't take the drug, his systolic blood pressure would be exactly $135.83$? Of course not! We have some uncertainty around these numbers, as well.  There are two main sources of uncertainty we have to take into account when making a prediction for Joe:

1. **Uncertainty in estimating the ATE**.  Especially with five subjects, we may not have a good estimate of the ATE.  With this miniscule sample, the uncertainty might be gigantic, like ($-25$, $13$), which would lead to a prediction for Joe of ($117, 155$).  As we get a larger sample size, this uncertainty decreases.
1. **Individual variation**.  Even if we have a perfect estimate of the *average* treatment effect, it still may be the case that the effect *for Joe* is higher or lower than average.  We can assume this away if we say that the treatment effect is constant for everyone, but that is not likely to be true in the real world -- and this source of uncertainty does not go away simply by collecting more observations.  So with a large sample, let's say that we calculated a confidence interval around the ATE of ($-6, -5.67$), leading to an interval for Joe's outcome under control of ($135.67, 136$).  The uncertainty due to individual variation may still be a great deal higher---say ($120, 152$).  These numbers are simply illustrative, but they highlight an important point: even if you have a good estimate of the ATE, you should still be much more uncertain about the causal effect for any *particular individual*.

### The assignment mechanism

<!-- DK: Everything is a missing data problem. -->

<!-- DK: Perhaps the best way to introduce this topic is to talk about just measuring a single outcome, like, say, income in the world. The Rubin Table for this just has a single column: income. If we were God, we could see all the values and calculate the the true median. (Even this is harder than it sounds because it is only for a given point in time. One day later some rows in the RT have disappeared (with someone's death) while other rows have been added, with someone turing 18.) But we don't get to see all the values. First, our sample, even if perfectly random, only includes X number of people, so we never had the chance to see 99.9% of the rows. Second, some of the sample refused to answer the question. They are also missing.  -->

<!-- How was it that some people (and not others) were sampled and some of the sample (but not others) did not report a number. In both cases, we want to know what process led to these outcomes. What was the "assignment mechanism" to sampling that led to Joe being included and Sarah not? For those included, what was the "assignment mechanism" to answering which led X to report her income and Y to not? -->

<!-- Once those ideas are explained, in the simplest possible context, we can move on to the normal assignment mechanism, which is a much trickier concept. -->


Is the difference in sample means between treated units and control units a good estimate of the average treatment effect?  That depends entirely on the method by which units are assigned treatment, which is called the **assignment mechanism**.

One such assignment mechanism is **randomization**. For each subject we could flip a coin to determine if she receives treatment. If we wanted five subjects to receive treatment, we could assign treatment to the first five names we pick out of a hat.

Randomized assignment is the best assignment mechanism for inferring average treatment effects because if the sample is large enough, the difference in sample means between treated and control units will be very close to the true average treatment effect in that sample.

However, when the sample size is small, our estimated average treatment effect may deviate considerably from the truth.  For example, the data in the last section were from just one possible random assignment, where Joe, Mary and James receive the treatment.  Here's another possible random assignment, where Joe, Mary and Sally receive the treatment:

```{r, echo = FALSE}
# First, we create a tibble with the values we want for the table

tibble(subject = c("Joe", "Mary", "Sally", "Bob", "James", "MEAN"),
       ytreat = c("130", "120", "100", "?", "?", "115"),
       ycontrol = c("?", "?", "?", "130", "120", "123"),
       ydiff = c("?", "?", "?", "?", "?", "-8.33")) %>%
  
  # Then, we use the gt function to make it pretty
  
  gt() %>%
  cols_label(subject = md("$$\\mathbf{Subject}$$"),
                ytreat = md("$$Y_t(u)$$"),
                ycontrol = md("$$Y_c(u)$$"),
                ydiff = md("$$Y_t(u) - Y_c(u)$$")) %>%
  tab_style(cell_borders(sides = "right"),
            location = cells_body(columns = vars(subject))) %>%
  tab_style(cell_text(weight = "bold"),
            location = cells_body(columns = vars(subject)))
```

Note that with one assignment, the estimate of the ATE is $-5.83$, while with another it is $-8.33$.  Assume that these data are the truth: 

```{r, echo = FALSE}
# First, we create a tibble with the values we want for the table

tibble(subject = c("Joe", "Mary", "Sally", "Bob", "James", "MEAN"),
       ytreat = c("130", "120", "100", "110", "115", "115"),
       ycontrol = c("115", "125", "130", "130", "120", "123"),
       ydiff = c("+15", "-5", "-25", "-20", "-5", "-8")) %>%
  
  # Then, we use the gt function to make it pretty
  
  gt() %>%
  cols_label(subject = md("$$\\mathbf{Subject}$$"),
                ytreat = md("$$Y_t(u)$$"),
                ycontrol = md("$$Y_c(u)$$"),
                ydiff = md("$$Y_t(u) - Y_c(u)$$")) %>%
  tab_style(cell_borders(sides = "right"),
            location = cells_body(columns = vars(subject))) %>%
  tab_style(cell_text(weight = "bold"),
            location = cells_body(columns = vars(subject)))
```

Thus, the true average treatment effect is $-8$.  However, our estimates of the average treatment effect vary because our sample is small and the responses have a large variance. If the sample were larger and the variance were less, the average treatment effect would be closer to the true average treatment effect regardless of the specific units randomly assigned to treatment.

### Permutation tests

<!-- AR: This section attempts to describe permutation tests in a non-frequentist
way; thus, there are no references to a "test statistic" or to the "sharp null." Worth keeping?
-->

Let's say we observed the following results, which we know came from a random treatment assignment:

```{r, echo = FALSE}
# First, we create a tibble with the values we want for the table

tibble(subject = c("Mary", "Sally", "Bob", "James", "MEAN"),
       ytreat = c("120", "100", "?", "?", "110"),
       ycontrol = c("?", "?", "130", "120", "125"),
       ydiff = c("?", "?", "?", "?", "-15")) %>%
  
  # Then, we use the gt function to make it pretty
  
  gt() %>%
  cols_label(subject = md("$$\\mathbf{Subject}$$"),
                ytreat = md("$$Y_t(u)$$"),
                ycontrol = md("$$Y_c(u)$$"),
                ydiff = md("$$Y_t(u) - Y_c(u)$$")) %>%
  tab_style(cell_borders(sides = "right"),
            location = cells_body(columns = vars(subject))) %>%
  tab_style(cell_text(weight = "bold"),
            location = cells_body(columns = vars(subject)))
```

The drug looks very useful!  However, how likely is it that we would observe these results if the drug did nothing at all?  That is, how confident should we be that the drug is actually helpful?

To answer this question, we can use a **permutation test**.^[See [this post](https://www.r-bloggers.com/what-is-a-permutation-test/) for a longer discussion.]  The intuition behind a permutation test is simple.  We observed four outcomes, two of which were assigned to treatment and two that were assigned to control.  To conduct a permutation test, we calculate our quantity of interest (here, the difference in means between treated and control) for every possible arrangement of the labels "treatment" and "control" across these four numbers.

This is easiest to understand visually:

```{r, echo = FALSE}
# First, we create a tibble with the values we want for the table

tibble(Permutation = c("#1", "#2", "#3", "#4", "#5", "#6"),
       `120` = c("T", "T", "T", "C", "C", "C"),
       `100` = c("T", "C", "C", "T", "T", "C"),
       `130` = c("C", "T", "C", "T", "C", "T"),
       `120 ` = c("C", "C", "T", "C", "T", "T"),
       ATE = c("-15", "+15", "+5", "-5", "-15", "+15")) %>%
       
  # Then, we use the gt function to make it pretty
  
  gt() %>%
  cols_label(Permutation = md("**Permutation**")) %>%
  tab_style(cell_borders(sides = "right"),
            location = cells_body(columns = vars(Permutation))) %>%
  tab_style(cell_text(weight = "bold"),
            location = cells_body(columns = vars(Permutation)))
```

Here, permutation #1 is what we actually observed, and permutations #2-6 are all the possible rearrangement of the two "treatment" labels and the two "control" labels.  What do we see?  4/6 (67%) of the permutations produce calculated ATEs are at least as large as the effect we actually observed ($-15$), and two of them are the opposite sign!  We therefore should not have much confidence from these data alone that the drug really lowers blood pressure, even though the effect size appeared large.  The moral of the story?  Don't conduct an experiment on only four people!

Let's say that we had a larger experiment.  We can no longer calculate the results of the permutation test by hand, since the number of possible permutations will quickly become very large.  However, you can use R to calculate far more permutations than you can do by hand.  Furthermore, if the number of permutations becomes too much even for your computer, you can take a random sample of all the possible permutations instead; if the random sample is large enough, this will give you a result very close to what you would get if you considered all the permutations.

<!-- DK: Ideally, we extend the Rubin Table until it is flexible enough to use in the context of regression. More complex statistical models, like regression, just provide a different way of imputing the values of those question marks. And we should show this clearly. -->

<!-- DK: Connect causal inference to prediction, even if only briefly.  -->

### Confounding and selection bias

With a large enough sample and randomized assignment, we can get a good estimate of the ATE.  Other assignment mechanisms, however, can make it difficult or impossible to engage in causal inference.

Why is that?  Consider the use of the *perfect doctor* as an assignment mechanism. The perfect doctor knows how each subject will respond to the drug or the control and assigns each subject to the treatment that will most benefit her. The perfect doctor knows this information about a sample of patients: 

```{r, echo = FALSE}
# First, we create a tibble with the values we want for the table

tibble(subject = c("Joe", "Mary", "Sally", "Bob", "James", "Susie", "MEAN"),
       ytreat = c("130", "120", "100", "115", "125", "135", "120"),
       ycontrol = c("115", "125", "150", "125", "130", "105", "125"),
       ydiff = c("+15", "-5", "-50", "-10", "-10", "+30", "-15")) %>%
  
  # Then, we use the gt function to make it pretty
  
  gt() %>%
  cols_label(subject = md("$$\\mathbf{Subject}$$"),
                ytreat = md("$$Y_t(u)$$"),
                ycontrol = md("$$Y_c(u)$$"),
                ydiff = md("$$Y_t(u) - Y_c(u)$$")) %>%
  tab_style(cell_borders(sides = "right"),
            location = cells_body(columns = vars(subject))) %>%
  tab_style(cell_text(weight = "bold"),
            location = cells_body(columns = vars(subject)))
```

Based on this knowledge she would make the following treatment assignments: 

```{r, echo = FALSE}
# First, we create a tibble with the values we want for the table

tibble(subject = c("Joe", "Mary", "Sally", "Bob", "James", "Susie", "MEAN"),
       ytreat = c("?", "120", "100", "115", "125", "?", "113.75"),
       ycontrol = c("115", "?", "?", "?", "?", "105", "110"),
       ydiff = c("?", "?", "?", "?", "?", "?", "+3.75")) %>%
  
  # Then, we use the gt function to make it pretty
  
  gt() %>%
  cols_label(subject = md("$$\\mathbf{Subject}$$"),
                ytreat = md("$$Y_t(u)$$"),
                ycontrol = md("$$Y_c(u)$$"),
                ydiff = md("$$Y_t(u) - Y_c(u)$$")) %>%
  tab_style(cell_borders(sides = "right"),
            location = cells_body(columns = vars(subject))) %>%
  tab_style(cell_text(weight = "bold"),
            location = cells_body(columns = vars(subject)))
```

The perfect doctor distorts the averages of both $Y_t(u)$ and $Y_c(u)$ by filtering out poor responses to both the treatment and control. The difference in means thus is distorted in a direction that depends on the details. For instance, a subject like Susie who is harmed by taking the drug would be assigned to the control group by the perfect doctor and thus the negative effect of the drug would be masked.

Therefore, the difference in means is no longer a good estimate of the ATE. In fact, in this case it has the wrong sign! This is not merely a consequence of our small sample: even if the perfect doctor saw a million patients, we could not get a good estimate of the ATE.

This is an extreme example of a problem called **selection bias**.  The perfect doctor is not choosing which patients are receiving the drug at random. Rather, the perfect doctor is making treatment decisions based directly on the patient's potential outcomes.  Whenever the assignment mechanism is correlated with the potential outcomes, we say that there is **confounding**.  Confounding is a problem, since it means that our simple estimate of the ATE is biased.

Thus, if you are interested in causal inference, randomized trials are the best approach. In many circumstances, however, randomized trials are not possible due to ethical or practical concerns. In such scenarios there is by necessity a non-random assignment mechanism.

For instance, let's say you were interested in the effect of college attendence on earnings. People are not randomly assigned to attend college. Rather, people may choose to attend college based on their financial situation, parents' education, and so on.  This can introduce confounding if the assignment mechanism affects future earnings.  For example, if people choose to go to college at higher rates when they are on career paths where a college degree is particularly beneficial  -- or, in RCM language, whose potential outcomes are on average higher under treatment -- that would introduce confounding.

Many statistical methods have been developed for causal inference when there is a non-random assignment mechanism, such as propensity score matching. These methods attempt to correct for the assignment mechanism by finding control units similar to treatment units. What you should not do is naively compare the sample means under treatment and control and assume that is a good estimate of the ATE.  Without randomization, this could be very misleading!

## Other issues with causal inference


### Stable unit treatment value assumption (SUTVA)

One important assumption for causal inference is that "the [potential outcome] observation on one unit should be unaffected by the particular assignment of treatments to the other units."^[Cox 1958, 2.4.] This is called the **Stable Unit Treatment Value Assumption** (**SUTVA**). 

In the context of our example, Joe's blood pressure should not depend on whether Mary receives the drug. But what if it does? Suppose that Joe and Mary live in the same house and Mary always cooks. The drug causes Mary to crave salty foods, so if she takes the drug she will cook with more salt than she would have otherwise. A high salt diet increases Joe's blood pressure. Therefore, his outcome will depend on both which treatment he received and which treatment Mary receives.

SUTVA violation makes causal inference more difficult. We can account for dependent observations by considering more treatments. We create 4 treatments by taking into account whether or not Mary receives treatment:

```{r, echo = FALSE}
# First, we create a tibble with the values we want for the table

tibble(subject = c("Joe"),
       ct = "140",
       tt = "130",
       cc = "125",
       tc = "120") %>%
  
  # Then, we use the gt function to make it pretty.
  
  gt() %>%
  cols_label(subject = md("**Subject**"),
             ct = html("Joe = c,<br />Mary = t"),
             tt = html("Joe = t,<br />Mary = t"),
             cc = html("Joe = c,<br />Mary = c"),
             tc = html("Joe = t,<br />Mary = c")) %>%
  tab_style(cell_borders(sides = "right"),
            location = cells_body(columns = vars(subject))) %>%
  tab_style(cell_text(weight = "bold"),
            location = cells_body(columns = vars(subject)))
```

Recall that a causal effect is defined as the difference between two potential outcomes. In this case, there are multiple causal effects because there are more than two potential outcomes:

- One is the causal effect of the drug on Joe when Mary receives treatment ($130-140$).
- Another is the causal effect of the drug on Joe when Mary does not receive treatment ($120-125$).
- A third is the causal effect of Mary's treatment on Joe when Joe is not treated ($140-125$). In other words, we can define a causal effect on Joe even in situation in which Joe's treatment is identical in both situations.

Note that Mary taking the drug has a larger causal effect on Joe than Joe himself taking the drug, and what's more, the effect is in the opposite dierction!

By considering more potential outcomes in this way, we can cause SUTVA to hold. However, if any units other than Joe are dependent on Mary, then we must consider further potential outcomes. The greater the number of dependent units, the more potential outcomes we must consider and the more complex the calculations become (consider an experiment with 20 different people, each of whose treatment status can affect outcomes for every one else). In order (easily) to estimate the causal effect of a single treatment relative to a control, SUTVA should hold. 

### Internal and external validity

If we have randomized assignment and a large sample, we can be confident that we have a good estimate of the average treatment effect *in that sample*.  We say that the experiment has high **internal validity**: the inferences we are making are likely to reflect the truth about that sample.

However, we may be interested in a population beyond our particular sample.  For example, consider the hypertension drug example discussed above.  Let's say that we ran a randomized trial on 10,000 patients, with 5,000 receiving the new drug and 5,000 in the contol group.  We estimate an average treatment effect of $-5$ units of systolic blood pressure.  Should we recommend this new drug?

The answer to that question depends in part on the **external validity** of the study.  Are the 10,000 people in the study similar to the people for whom we will be making recommendations?  Let's say that the 10,000 study participants were chosen because they all lived near the location that the study took place. That's one example of **selection bias**, where the sample is not a random sample of the population in which we are interested.  Why is that a problem?  Those people may differ systematically from other people, including in ways that affect their response to blood pressure medicine. For example, what if the people in this area are younger on average than most people who take blood pressure medicine?

This concern can be expressed in terms of the assignment mechanism.  People who live far away from the location of the study have a 0% chance of receiving the treatment.  Thus, the study *can't* directly speak to whether the treatment is effective for them; the only way we can make such claims is by making additional assumptions, such as that the non-participants will respond the same on average to the drug as the participants.

The circumstances of the experiment may also affect the external validity of the study.  Perhaps the study participants were all eating a diet that was controlled by the experimenters.  Then, while we have variation in one aspect of the treatment (whether the participants received the blood pressure medicine), we don't have variation in another (diet).  It may be that the blood pressure medicine works when on the controlled diet but doesn't work otherwise.

When dealing with human subjects, there is a particular concern regarding external validity: the [**Hawthorne effect**](https://en.wikipedia.org/wiki/Hawthorne_effect).  When human subjects know that they are part of an experiment, they may change their behavior.

This can lead to a version of the above concern.  Suppose again that the blood pressure drug works when combined with a controlled diet, say a low-salt diet, but it doesn't work for those who have a high-salt diet.  If the subjects in the experiment are paying greater attention to their blood pressure than they otherwise would and thus are eating lower sodium diets than other people, one could make the mistaken inference that it was the blood pressure drug alone that led to a decrease in blood pressure, and not the combination of the drug and the experimental subjects' changing behavior.

## Conclusion

The causal effect of a treatment on a single unit at a point in time is the difference between the outcome variable with the treatment and without the treatment. The Fundamental Problem of Causal Inference is that it is impossible to observe the causal effect on a single unit. You either take the blood pressure drug or you don't. As a consequence, assumptions must be made in order to estimate the missing counterfactuals.

## References

<!-- AR: Need to integrate this with the rest of the book -->

Cox, D. R. (1958). *Planning of Experiments.* Wiley.

Holland, Paul W. (1986). "Statistics and Causal Inference." *J. Amer. Statist. Assoc.* 81 (396): 945960. doi:10.1080/01621459.1986.10478354.