---
output_yaml:
  - output.yml
---

# Simpson's Paradox {-}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tufte)
```


<!-- create full width div -->

::: {.fullwidth} 

## **What is Simpson's Paradox? The Basics:**

To navigate the tutorial, click on the box below and use your left and right arrow keys to explore the basic ideas of [Simpson’s Paradox](https://en.wikipedia.org/wiki/Simpson%27s_paradox).

<!-- end full width div -->

:::

```{marginfigure} 

Simpson’s Paradox is a statistical phenomenon in which trends for data in the aggregate either disappear or are reversed when the data are broken down into groups.

```

```{r embed_basic_example_flipbook, echo = FALSE, fig.align='center'}

# embed the flipbook linked from the existing folder into the html page here:

knitr::include_url('simpsons-paradox/basic_simpsons_flipbook.html')

```

<!-- create full width div -->

::: {.fullwidth}

## **Building Understanding: Examples of Simpson's Paradox**

Simpson's Paradox is most easily understood through examples. The following example will help us to build our understanding of this phenomenon and enhance our ability to recognize and understand it in future examples. 

### **The Credit Card Debt and Income Example**

In this example, the same credit data can be used to tell contradictory stories. When looking at the relationship between debt and income alone, the relationship appears to be positive. However, when we take into account credit limit as well, the relationship between income and debt becomes negative. How can the data suggest that the relationship between debt and income is both positive *and* negative?

<!-- end full width div -->

:::

```{marginfigure} 

Income and credit debt are positively related in the aggregate but negatively related when we control for an individual's credit limit.

```


```{r embed_credit_flipbook, echo = FALSE, fig.align='center'}

# embed the flipbook linked from the existing folder into the html page here:

knitr::include_url('simpsons-paradox/credit_example_flipbook1.html')

```

<!-- create full width div -->

::: {.fullwidth}

Here, credit limit is a *confounding variable* of the relationship between credit card debt and income. Thus, credit limit needs to be accounted for in any appropriate model for the relationship between debt and income. Understanding the role of the confounding variable of credit limit helps us resolve Simpson's Paradox in this case.

## **When Should We Focus on Relationships in the Aggregate v. Within Groups?**

The tricky thing about Simpson's Paradox is that it can mean that vastly different interpretations can be taken from the same data. But which interpretation is right? Should we always look at relationships on the aggregate level, or should we break them down according to all the relevant groups and confounding variables? Unfortunately, the answer to this important questions is that it depends. It depends largely on the context of the problem, the question we are trying to answer, and what data is available.

For example, if we are aware that a confounding variable in our analysis has been ignored and we have the relevant data on it, then we should include that variable in our analysis and break it down from the aggregate case. If we were trying to predict the debt of a new individual with a certain income and credit limit, for example, it would be best to predict her debt using the model that takes credit limit into account. If we did not, our prediction would likely be much less accurate. 
 
However, if we do not know what the confounding variable in our problem is - for example, if we did not know that credit limit is negatively associated with debt - or if we do not have data on the confounding variable, then it is best and often only possible to look at trends and relationships in the aggregate. 

## **So What?**

Simpson's Paradox is a phenomenon that is frequently encountered in social science and medical science research. It highlights the importance of carefully considering model specification and potential confounding variables, because it shows how we can draw contradictory conclusions from the same data if we are not careful. Above all, Simpson's Paradox is a great example of why it is critical that we, as data scientists, always look at our work with a skeptical eye and why we must be hesitant to trust what may seem like obvious conclusions, even if they are supported by initial analysis.

Importantly, it is crucial to remember that no data is purely objective. The data and relationships that we see from models and plots are only one (often greatly simplified) abstraction of the phenomena that actually occur in the real world. In order to be true data scientists, we must always ask oursleves *why* the data look the way they do and what underlying mechanisms may be driving any relationships we observe. Especially when interpreting the results of our models, it is key to ask ourselves about the broader context of the data, the data generating process, and *what might be missing*. In doing this, we are less likely to draw inappropriate conclusions. 

<!-- end full width div -->

:::
