---
title: "Visualization"
output: learnr::tutorial
runtime: shiny_prerendered
---


<!-- DK: I think that the first 20 to 40 questions should, perhaps counterintuitively, have nothing to do with graphics. Graphics are hard! These are the first R commands that they are typing. We want them to be easy. Recall the Basic programming concepts section. Right now that goes fast. In fact, it goes way too fast for anyone who has not programmed before. Let's slow it down in the chapter. And slow it down here. Ask lots of simple questions using those basic tools. For example, there should be at least two or three questions on using `%in%` correctly. And two or three on `ifelse()`. And also questions about all the common commands that we use all the time by themselves: `glimps()`, `View()`, `summary()` and so on. -->

<!-- Again, these will be easy questions: something like "Use `glimpse()` to examine the mtcars data set." With "`glimpse(mtcars)` being the answer. And then, perhaps, a multiple choice question which asks for the type of one of the variables in `mtcars`. That is, obviously, very simple! And that is OK! We just want to force students to, at least one time, type `glimpse(mtcars)` and look at the output. We also want to show them all the handy arguments we actually use. Example: I use `print(x)` to look at x, but also `print(x, n = 50)` to look at more than the default 10 observations and `print(x, width = 200)` to look at more of the variables than the default gives you. We should probably survey the entire group for examples of functions which they use all the time. We don't want weird or hard functions, or anything that is mainly used with pipes. We want commonly used things. And we want to questions to parallel the order we present them in the book. -->

<!-- Some other key commands to cover in the book and in the tutorial: `install.packages()`, `library()`, `<-` and . . . should also look in chapters 2, 3, and 4 to see what might be useful to include now. Maybe `case_when`? Note that this would mean that we should start with `install.packages("tidyverse")` and `library(tidyverse)`. -->

<!-- Also, how to look up help, like `?case_when`. Need to teach them how to use functions, including setting arguments. Maybe `seq()` is a good example, even though we don't use it much later? Or use `head()` and `tail()` for this at first, since you can introduce `n` as the first argument to teach them about? -->

<!-- Maybe teach them to top 4 short cut keys? Especially how to comment a block of code and then reflow it. Does this "work" within a tutorial? Like, we provide a poorly formatted comments and they need to reformat it? -->

<!-- But we don't teach the pipe until the next chapter. -->


<!-- Template for tutorials to add questions from Chapter 1 to -->

<!-- Installed the learnr package, created a new R Markdown file by selecting the "From Template" tab and the "Interactive Tutorials" template -->

```{r setup, include=FALSE}
library(learnr)

# Install the gradethis package with remotes::install_github("rstudio-education/gradethis")
# Made sure you have the latest versions of R and RStudio before installing

library(gradethis) 

knitr::opts_chunk$set(echo = FALSE, exercise.checker = grade_learnr)

# tutorial_event_recorder <- function(tutorial_id, tutorial_version, user_id, 
#                                     event, data) {
#   cat(tutorial_id, " (", tutorial_version, "): ", user_id , "\n", sep = "")
#   cat("event: ", event, "\n", sep = "")
# }

# options(tutorial.storage = filesystem_storage(dir = "./storage", compress = TRUE))
```

## What are R and RStudio?

```{r quiz-1}
quiz(
  question("What are R and RStudio?",
             answer("R is a programming language and RStudio is an integrated development environment (IDE)", 
                    correct = TRUE),
             answer("RStudio is a programming language and R is an IDE"),
             answer("R and RStudio are both programming languages"),
             answer("R and RStudio are both IDEs"),
             allow_retry = TRUE
    ),
  question("In what order do you install R and RStudio?",
             answer("First R then RStudio", correct = TRUE),
             answer("First RStudio then R"),
             allow_retry = TRUE
    )
)
```

## How Do I Code in R?

```{r quiz-2}
quiz(
  question("Which of the following is NOT the name of a data type in R?",
           answer("Logicals"),
           answer("Integers"),
           answer("Characters"),
           answer("Strings", correct = TRUE),
           allow_retry = TRUE
  ),
  question("What is the operator used in R to test for inclusion?",
           answer("'in'"),
           answer("'%in%'", correct = TRUE),
           answer("'include'"),
           answer("'IN'"),
           allow_retry = TRUE
  ),
  question("What is the symbol used to bring up a help file?",
           answer("*"),
           answer("~"),
           answer("!"),
           answer("?", correct = TRUE),
           allow_retry = TRUE
  ),
  question("True / False: Seeing red text in the console is always bad because it denotes an error.",
           answer("True"),
           answer("False", correct = TRUE, message = "R will show red text in the console pane for errors, warnings, and messages. The only situation in which the code will not run is if there is an error."),
           incorrect = "R will show red text in the console pane for errors, warnings, and messages. The only situation in which the code will not run is if there is an error.",
           allow_retry = TRUE
  )
)
```

## What are R packages?

```{r quiz-3}
quiz(
  question("Which of the following is NOT true about R packages?",
           answer("R packages extend the functionality of R by providing additional functions, data, and documentation"),
           answer("If comparing R and R packages to a phone, R would be the apps you can download on the phone and R packages would be the phone", correct = TRUE),
           answer("R packages are written by a worldwide community of R users"),
           answer("R packages can be downloaded for free from the Internet"),
           allow_retry = TRUE
  ),
  question("Which command corresponds to loading a package in R?",
           answer("install.packages(“package”)"),
           answer("library(package)", correct = TRUE),
           allow_retry = TRUE
  )
)
```

*Load the `ggplot2` package.*
```{r exercise-1, exercise = TRUE, exercise.lines = 3}

```

```{r exercise-1-solution}
library(ggplot2)
```

```{r exercise-1-check}
grade_code()
```

## Explore your first datasets

*These questions will use the `trains` dataset within the `PPBDS.data` package. The package should be already installed on your computer. If not, run `remotes::install_github(“davidkane9/PPBDS.data”)` in the RStudio console pane.*

*Load the `PPBDS.data` package to look at the `trains` dataset.*
```{r exercise-2, exercise = TRUE, exercise.lines = 5}
---------
  
trains
```

```{r exercise-2-solution}
library(PPBDS.data)

trains
```

```{r exercise-2-check}
grade_code()
```

<!-- *Use the View() function to look at the trains dataset.* -->

<!-- JSE - View() doesn't work that well in the tutorials, opens up a new tab with the data that looks kind of ugly -->

<!-- ```{r exercise-3, exercise = TRUE, exercise.lines = 5} -->
<!-- --------- -->

<!-- ``` -->

<!-- ```{r exercise-3-solution} -->
<!-- View(trains) -->
<!-- ``` -->

*Use the `glimpse()` function to look at the `trains` dataset. We already loaded the `dplyr` and `PPBDS.data` packages.*
```{r exercise-4, exercise = TRUE, exercise.lines = 5}
library(dplyr)
library(PPBDS.data)

---------
```

```{r exercise-4-solution}
library(dplyr)
library(PPBDS.data)

glimpse(trains)
```

```{r quiz-4}
quiz(
  question("What is the data type of the `liberal` variable?",
           answer("character"),
           answer("logical", correct = TRUE),
           answer("double"),
           answer("factor"),
           allow_retry = TRUE
  ),
  question("How many rows does the `train` dataset have?",
           answer("8"),
           answer("15"),
           answer("115", correct = TRUE),
           allow_retry = TRUE
  )
)
```

```{r exercise-4-check}
grade_code()
```

*Extract the `income` variable in the `trains` dataset using the `$` operator.*
```{r exercise-5, exercise = TRUE, exercise.lines = 5}
library(PPBDS.data)

---------
```

```{r exercise-5-solution}
library(PPBDS.data)

trains$income
```

```{r exercise-5-check}
grade_code()
```

*Look at the `qscores` dataset using `glimpse()`. We have loaded the `PPBDS.data` and the `dplyr` packages.*
```{r qscores-setup, include = FALSE}
library(PPBDS.data)
library(dplyr)
```

```{r qscores, echo = TRUE}
glimpse(qscores)
```

```{r quiz-5}
quiz(
  question("Which of the following is NOT an identification variable?",
           answer("`course_name`"),
           answer("`department`", correct = TRUE),
           answer("`workload`", correct = TRUE),
           answer("`course_number`"),
           allow_retry = TRUE
  )
)
```

## Components of a graphic

```{r quiz-6}
quiz(
  question("What are the three essential components of a graphic?",
           answer("the dataset containing the variables in question", correct = TRUE),
           answer("the geometric object we can observe in a plot", correct = TRUE),
           answer("axes labels on a plot"),
           answer("the aesthetic attributes", correct = TRUE),
           allow_retry = TRUE
  ),
  question("What are the two important arguments that we need to provide the `ggplot()` function?",
           answer("`data` and `mapping`", correct = TRUE),
           answer("`data` and `aesthetics`"),
           answer("`data` and `layers`"),
           allow_retry = TRUE
  )
)
```

*Look at this scatterplot made using the `trains` dataset. We have loaded the `PPBDS.data` and the `ggplot2` packages.*

*Higher numbers mean more conservative attitudes toward immigration-related policies on a 3 to 15 scale. Individuals with a treatment value of "Treated" were exposed to two Spanish-speakers on their regular commute. "Control" individuals were not.*

```{r trains}
library(PPBDS.data)
library(ggplot2)

ggplot(data = trains, mapping = aes(x = att_start, y = att_end, color = treatment)) +
  geom_point() + facet_wrap(~party)
```

```{r quiz-7}
quiz(
  question("Which `data` variable is mapped to the `x`-position `aes`thetic of the points?",
           answer("`att_start`", correct = TRUE),
           answer("`att_end`"),
           answer("`treatment`"),
           allow_retry = TRUE
  ),
  question("Which `data` variable is mapped to the `y`-position `aes`thetic of the points?",
           answer("`att_start`"),
           answer("`att_end`", correct = TRUE),
           answer("`treatment`"),
           allow_retry = TRUE
  ),
  question("Which `data` variable is mapped to the `color` `aes`thetic of the points?",
           answer("`att_start`"),
           answer("`att_end`"),
           answer("`treatment`", correct = TRUE),
           allow_retry = TRUE
  )
)
```

## Scatterplots

*Let's make a scatterplot using the `qscores` dataset. Load the `PPBDS.data` package and look at the `qscores` dataset by simply typing the name of the dataset.*

```{r exercise-6, exercise = TRUE, exercise.lines = 5}
---------

---------
```

```{r exercise-6-solution}
library(PPBDS.data)

qscores
```

```{r exercise-6-check}
grade_code()
```

*Fill in the `ggplot()` function to create a scatterplot with `overall` on the x-axis and `workload` on the y-axis. Is there a relationship between the overall rating of a Harvard course versus the average hours per week of workload?*

```{r exercise-7, exercise = TRUE, exercise.lines = 5}
library(ggplot2)

ggplot(data = qscores, mapping = aes(x = ---------, y = ---------)) +
  geom_point()
```

```{r exercise-7-solution}
ggplot(data = qscores, mapping = aes(x = overall, y = workload)) + 
  geom_point()
```

```{r exercise-7-check}
grade_code()
```

*Now we want to add a `size` `aes`thetic based on the number of students enrolled in each course. Add `size = enrollment` to the other `aes`thetics of the points. Do larger classes have a higher average overall course rating? What about higher workload?*

```{r exercise-8, exercise = TRUE, exercise.lines = 5}
ggplot(data = qscores, mapping = aes(x = overall, y = workload, ---------)) +
  geom_point()
```

```{r exercise-8-solution}
ggplot(data = qscores, mapping = aes(x = overall, y = workload, size = enrollment)) +
  geom_point()
```

```{r exercise-8-check}
grade_code()
```

*Change the transparency of the points by setting `alpha = 0.2` within `geom_point()`.*

```{r exercise-9, exercise = TRUE, exercise.lines = 5}
ggplot(data = qscores, mapping = aes(x = overall, y = workload, size = enrollment)) +
  geom_point(----------)
```

```{r exercise-9-solution}
ggplot(data = qscores, mapping = aes(x = overall, y = workload, size = enrollment)) +
  geom_point(alpha = 0.2)
```

```{r exercise-9-check}
grade_code()
```

*Replace `geom_point()` with `geom_jitter()` to create a jittered scatterplot.*

```{r exercise-10, exercise = TRUE, exercise.lines = 5}
ggplot(data = qscores, mapping = aes(x = overall, y = workload)) +
  geom_point()
```

```{r exercise-10-solution}
ggplot(data = qscores, mapping = aes(x = overall, y = workload)) +
  geom_jitter()
```

```{r exercise-10-check}
grade_code()
```

## Linegraphs

```{r linegraphs-setup, include = FALSE}
library(PPBDS.data)
library(tidyverse)

cces_data <- cces %>%
  filter(approval_gov %in% c("Approve / Somewhat Approve", "Disapprove / Somewhat Disapprove")) %>%
  group_by(year, approval_gov) %>%
  summarize(avg_age = mean(age))
```

*The `cces` dataset in the `PPBDS.data` package has been changed to show the average age of voters per year based on their approval of the president that year. The `cces` dataset was manipulated using the `filter()`, `group_by`, and `summarize()` functions within the `tidyverse` package, which you will learn how to use in the next chapter.*

```{r gov_qscores, echo = TRUE}
cces_data
```

*Fill in the `ggplot()` function to create a linegraph with `year` on the x-axis, `avg_age` on the y-axis, and `approval_ch` as the color aesthetic.*

```{r exercise-11, exercise = TRUE, exercise.lines = 5}
library(ggplot2)

ggplot(data = cces_data, mapping = aes(x = ---------, y = ---------, color = ----------)) +
  geom_line()
```

```{r exercise-11-solution}
ggplot(data = cces_data, mapping = aes(x = year, y = avg_age, color = approval_ch)) + 
  geom_line()
```

```{r exercise-11-check}
grade_code()
```

```{r quiz-8}
quiz(
  question("True / False: Linegraphs and scatterplots both display the relationship between two numerical variables.",
           answer("True", correct = TRUE),
           answer("False"),
           allow_retry = TRUE),
  question("True / False: It is preferred to use scatterplots over linegraphs when the variable on the x-axis has an inherent ordering, like in a time series plot.",
           answer("True"),
           answer("False", correct = TRUE),
           allow_retry = TRUE)
)
```

## Histograms

```{r quiz-9}
quiz(
  question("A histogram is:",
           answer("A plot that displays the relationship between two numerical variables"),
           answer("A plot that visualizes the distribution of a numerical variable", correct = TRUE),
           answer("A plot that visualizes the distribution of a categorical variable"),
           allow_retry = TRUE)
)
```

*Using the `qscores` dataset and `ggplot()`, make a histogram where `overall` is on the x-axis to see the distribution of Harvard course ratings.*

```{r exercise-12, exercise = TRUE, exercise.lines = 5}
library(PPBDS.data)
library(ggplot2)

ggplot(data = qscores, mapping = aes(x = ---------)) + geom_histogram()
```

```{r exercise-12-solution}
library(PPBDS.data)
library(ggplot2)

ggplot(data = qscores, mapping = aes(x = overall)) + geom_histogram()
```

```{r exercise-12-check}
grade_code()
```

```{r quiz-10}
quiz(
  question_text("What is the default value of bins for this histogram?",
                answer("30", correct = TRUE), 
                allow_retry = TRUE)
)
```

*Add white vertical borders demarcating the bins by adding `color = "powderblue"` and `fill = "violet"` arguments to `geom_histogram()`.*

```{r exercise-13, exercise = TRUE, exercise.lines = 5}
ggplot(data = qscores, mapping = aes(x = overall)) +
  geom_histogram(color = --------, fill = --------)
```

```{r exercise-13-solution}
ggplot(data = qscores, mapping = aes(x = overall)) + 
  geom_histogram(color = "powderblue", fill = "violet")
```

```{r exercise-13-check}
grade_code()
```

*Specify the number of bins to be 10 via the `bins` argument in `geom_histogram()`.*

```{r exercise-14, exercise = TRUE, exercise.lines = 5}
ggplot(data = qscores, mapping = aes(x = overall)) +
  geom_histogram(bins = --------, color = "white")
```

```{r exercise-14-solution}
ggplot(data = qscores, mapping = aes(x = overall)) +
  geom_histogram(bins = 10, color = "white")
```

```{r exercise-14-check}
grade_code()
```

*Specify the width of the bins to be 1 via the `binwidth` argument in `geom_histogram()`.*

```{r exercise-15, exercise = TRUE, exercise.lines = 5}
ggplot(data = qscores, mapping = aes(x = overall)) +
  geom_histogram(binwidth = ---------, color = "white")
```

```{r exercise-15-solution}
ggplot(data = qscores, mapping = aes(x = overall)) +
  geom_histogram(binwidth = 1, color = "white")
```

```{r exercise-15-check}
grade_code()
```

## Boxplots

```{r quiz-11}
quiz(
  question("Which of the following are included in the five-number summary of a numerical variable used to construct a boxplot?",
           answer("Minimum", correct = TRUE),
           answer("First quartile", correct = TRUE),
           answer("Median", correct = TRUE),
           answer("Mean"),
           answer("Third quartile", correct = TRUE),
           answer("Maximum", correct = TRUE)
  )
)
```

*Let's keep playing with the `qscores` dataset in the `PPBDS.data` package. We have filtered the dataset using the `tidyverse` package based on the five departments with the classes with the most enrollment: ECON, COMPSCI, STAT, GOV, and MATH. (You will learn how to do this in the next chapter). The filtered dataset is called `top_five_qscores`.*

```{r boxplot-setup, include = FALSE}
library(tidyverse)
library(PPBDS.data)

top_five_enrolled <- qscores %>% 
  group_by(department) %>%
  summarize(enrollment = sum(enrollment)) %>% 
  arrange(desc(enrollment)) %>%
  pull(department)

top_five_qscores <- qscores %>% filter(department %in% top_five_enrolled[0:5])
```

```{r boxplot-top-five-qscores, echo = TRUE}
top_five_qscores
```

*Using the `top_five_qscores` dataset and `ggplot()`, make a boxplot where `department` is on the x-axis and `overall` is on the y-axis.*

```{r exercise-16, exercise = TRUE, exercise.lines = 5}
library(ggplot2)

ggplot(data = top_five_qscores, mapping = aes(x = ---------, y = ---------)) + geom_boxplot()
```

```{r exercise-16-solution}
library(ggplot2)

ggplot(data = top_five_qscores, mapping = aes(x = department, y = overall)) + geom_boxplot()
```

```{r exercise-16-check}
grade_code()
```

```{r quiz-12}
quiz(
  question("Which department has the highest median course rating?",
           answer("COMPSCI"),
           answer("ECON"),
           answer("GOV"),
           answer("MATH", correct = TRUE),
           answer("STAT")
  ),
  question("Which department has outliers?",
           answer("COMPSCI"),
           answer("ECON"),
           answer("GOV"),
           answer("MATH", correct = TRUE),
           answer("STAT")
  ),
  question("Which department has the largest range of course ratings?",
           answer("COMPSCI"),
           answer("ECON"),
           answer("GOV", correct = TRUE),
           answer("MATH"),
           answer("STAT")
  )
)
```

## Barplots

```{r quiz-13}
quiz(
  question("A barplot is:",
           answer("A plot that displays the relationship between two numerical variables"),
           answer("A plot that visualizes the distribution of a numerical variable"),
           answer("A plot that visualizes the distribution of a categorical variable",
                  correct = TRUE),
           allow_retry = TRUE)
)
```

*We can now use a barplot to plot how many students are enrolled in classes per department. We will use the `top_five_qscores` dataset again to look at the five largest departments.*

```{r barplot-setup, include = FALSE}
library(tidyverse)
library(PPBDS.data)

top_five_enrolled <- qscores %>% 
  group_by(department) %>%
  summarize(enrollment = sum(enrollment)) %>% 
  arrange(desc(enrollment)) %>%
  pull(department)

top_five_qscores <- qscores %>% filter(department %in% top_five_enrolled[0:5])
```

```{r barplot-top-five-qscores, echo = TRUE}
top_five_qscores
```

*Use `geom_bar()` to plot `department` on the x-axis to see the number of classes in each department.*

```{r exercise-17, exercise = TRUE, exercise.lines = 5}
library(ggplot2)

ggplot(data = top_five_qscores, mapping = aes(x = ---------)) +
  ---------
```

```{r exercise-17-solution}
library(ggplot2)

ggplot(data = top_five_qscores, mapping = aes(x = department)) +
  geom_bar()
```

```{r exercise-17-check}
grade_code()
```

*We can then count the number of classes per department in order to use the `geom_col()` layer instead. We do this using the `group_by()` and `count()` functions in the `tidyverse` package, which you will learn how to use in the next chapter. Our new data frame is stored as `top_five_qscores_counted`.*

```{r classes-counted-setup, include = FALSE}
library(tidyverse)

top_five_qscores_counted <- top_five_qscores %>% group_by(department) %>% count() %>% rename(count = n)
```

```{r avg-workload-setup, echo = TRUE}
top_five_qscores_counted
```

*Use `geom_col()` to plot `department` on the x-axis and `count` on the y-axis to see the number of classes in each department.*

```{r exercise-18, exercise = TRUE, exercise.lines = 5}
ggplot(data = top_five_qscores_counted, mapping = aes(x = ---------, y = ---------)) +
  ---------
```

```{r exercise-18-solution}
ggplot(data = top_five_qscores_counted, mapping = aes(x = department, y = count)) +
  geom_col()
```

```{r exercise-18-check}
grade_code()
```

```{r quiz-14}
quiz(
  question("What's the difference between `geom_bar()` and `geom_col()`?",
           answer("`geom_bar()` is used to represent pre-counted data while `geom_col() is not."),
           answer("`geom_col()` is used to represent pre-counted data while `geom_bar() is not.", 
                  correct = TRUE),
           answer("There is no difference between the two layers.")
  )
)
```

*Let's go back to the `trains` dataset in the `PPBDS.data` package. Use `geom_bar()` to plot `treatment` on the x-axis.*

```{r exercise-19, exercise = TRUE, exercise.lines = 5}
library(PPBDS.data)
library(ggplot2)
 
ggplot(data = trains, mapping = aes(x = ---------)) + ---------
```

```{r exercise-19-solution}
library(PPBDS.data)
library(ggplot2)
 
ggplot(data = trains, mapping = aes(x = treatment)) + geom_bar()
```

```{r exercise-19-check}
grade_code()
```

*We can now map the additional variable `party` by adding a `fill = party` inside the `aes()` aesthetic mapping.*

```{r exercise-20, exercise = TRUE, exercise.lines = 5}
ggplot(data = trains, mapping = aes(x = treatment, fill = --------)) + geom_bar()
```

```{r exercise-20-solution}
ggplot(data = trains, mapping = aes(x = treatment, fill = party)) + geom_bar()
```

```{r exercise-20-check}
grade_code()
```

*Change `fill = party` to `color = party`. What happens? *

```{r exercise-21, exercise = TRUE, exercise.lines = 5}
ggplot(data = trains, mapping = aes(x = treatment, fill = party)) + geom_bar()
```

```{r exercise-21-solution}
ggplot(data = trains, mapping = aes(x = treatment, color = party)) + geom_bar()
```

```{r exercise-21-check}
grade_code()
```

*Create a side-by-side barplot by adding `position = "dodge"` to `geom_bar()`.*

```{r exercise-22, exercise = TRUE, exercise.lines = 5}
ggplot(data = trains, mapping = aes(x = treatment, fill = party)) + geom_bar(---------)
```

```{r exercise-22-solution}
ggplot(data = trains, mapping = aes(x = treatment, fill = party)) + geom_bar(position = "dodge")
```

```{r exercise-22-check}
grade_code()
```

*Instead of mapping `party` to `fill`, we can include it as the variable to create small multiples of the plot across the levels of `party` using `facet_wrap(~ party)`. Add this layer to the provided plot in order to create a faceted barplot.*

```{r exercise-23, exercise = TRUE, exercise.lines = 5}
ggplot(data = trains, mapping = aes(x = treatment)) + geom_bar() + ---------
```

```{r exercise-23-solution}
ggplot(data = trains, mapping = aes(x = treatment)) + geom_bar() + facet_wrap(~ party)
```

```{r exercise-23-check}
grade_code()
```

<!-- ## Adding Layers to a Plot -->

<!-- Need to add: faceting, position, scales, coordinate systems, statistical transformations, adding titles and labels to a graph -->


